Cell-type-specific fluorescent tagging of endogenous target proteins reveals synaptic enrichment and dynamic regulations of dopamine receptors

Version 1

Abstract

Abstract

Dopamine can play opposing physiological roles depending on the receptor subtype. In the fruit fly Drosophila melanogaster, Dop1R1 and Dop2R encode the D1- and D2-like receptors, respectively, and are reported to oppositely regulate intracellular cAMP levels. Here, we profiled the expression and subcellular localization of endogenous Dop1R1 and Dop2R in specific cell types in the mushroom body circuit. For cell-type-specific visualization of endogenous proteins, we employed reconstitution of split-GFP tagged to the receptor proteins. We detected dopamine receptors at both presynaptic and postsynaptic sites in multiple cell types. Quantitative analysis revealed enrichment around the active zones, particularly for Dop2R. The presynaptic localization of Dop1R1 and Dop2R in dopamine neurons suggests dual feedback regulation as autoreceptors. Furthermore, we discovered a starvation-dependent, bidirectional modulation of the presynaptic receptor expression in the PAM and PPL1 clusters, two distinct subsets of dopamine neurons, suggesting regulation of appetitive behaviors. Our results highlight the significance of the co-expression of the two antagonizing dopamine receptors in the spatial and conditional regulation of dopamine responses in neurons.

Introduction

Neurotransmitters typically have multiple cognate receptors, and they may recruit different second messenger systems. Therefore, the expression and localization of receptor subtypes are critical for determining cellular responses to neurotransmitter inputs. The dopaminergic system offers an ideal in-vivo study case to this end, as it regulates a wide array of physiological functions through combinations of different receptor subtypes. In mammals, D1-like receptors are coupled to Gαs, thereby activating adenylate cyclase upon ligand binding, whereas Gαi-coupled D2-like receptors inhibit cyclase activity (Beaulieu & Gainetdinov, 2011). Four dopamine receptors have been identified in Drosophila: Dop1R1, Dop1R2, Dop2R, and DopEcR (K. A. Han et al., 1996; Hearn et al., 2002; Srivastava et al., 2005; Sugamori et al., 1995). Dop1R1 and Dop2R correspond to the D1- and D2-like receptors, respectively (Hearn et al., 2002; Sugamori et al., 1995). Dop1R2 and DopEcR are invertebrate-specific and have been reported to recruit different second messenger systems (K. A. Han et al., 1996; Srivastava et al., 2005). Intriguingly, recent data of single-cell RNA-seq and transgenic expression profiling revealed that the expression of these dopamine receptors is highly overlapping in the fly brain (Croset et al., 2018; Davie et al., 2018; Dolan et al., 2019; Kondo et al., 2020), unlike the spatially segregated expression of the D1- and D2-like receptors in vertebrate brains (Gerfen & Surmeier, 2011). Considering the opposing physiological roles of the Dop1R1 and Dop2R, their protein localization especially in those cells where they are coexpressed should be critical to determine the responses to the dopamine inputs.

Drosophila mushroom bodies (MB) have long served as a unique dopaminergic circuit model to study adaptive behaviors, such as associative memory. MB-projecting neurons and their connections have been systematically described at both mesoscopic and ultrastructural resolutions (Aso, Hattori, et al., 2014; Li et al., 2020; Takemura et al., 2017; Tanaka et al., 2008). Kenyon Cells (KCs), the major MB intrinsic neurons, encode a variety of sensory information (Honegger et al., 2011; Turner et al., 2008; Vogt et al., 2014). Intriguingly, each KC receives synaptic inputs from different dopaminergic projections in multiple spatially segmented compartments along its axon in the MB lobe (Aso, Hattori, et al., 2014; Tanaka et al., 2008). MB-projecting dopamine neurons (DANs) originate from the three cell clusters (Mao & Davis, 2009; Nässel & Elekes, 1992). DANs in the protocerebral posterior lateral 1 (PPL1) cluster project to the vertical lobes and the peduncle of the MB, and they control different aspects of associative memory (Aso et al., 2010, 2012; Aso & Rubin, 2016; Claridge-Chang et al., 2009; Krashes et al., 2009; Mao & Davis, 2009; Masek et al., 2015; Riemensperger et al., 2005; Takemura et al., 2017; Vogt et al., 2014). The protocerebral anterior medial (PAM) cluster, the largest DAN cluster, mostly projects to the medial lobes of the MB, and many PAM neurons are involved in reward processing (Burke et al., 2012; Felsenberg et al., 2018; Huetteroth et al., 2015; Ichinose et al., 2021; Lin et al., 2014; Liu et al., 2012; Yamagata et al., 2015, 2016). DANs in PPL2ab project to the MB calyx and control the conditioned odor response (Boto et al., 2019). In addition to the variety of the dopamine sources, KCs express all four dopamine receptor subtypes (Deng et al., 2019; Kondo et al., 2020). Given the multitude of modulatory effects of dopamine in the MB (Berry et al., 2018; Cohn et al., 2015; Handler et al., 2019), receptor localization in KCs provides important information for interpreting such functional diversity.

The projections and synapses of the neurons in the MB circuit are tightly intertwined (Li et al., 2020; Takemura et al., 2017). Therefore, conventional immunohistochemical approaches using light microscopy do not allow identification of cells from which immunoreactive signals originate. Precise determination of their subcellular localization requires visualization of the proteins of interest only in the target MB neurons (Fendl et al., 2020). Employing the CRISPR/Cas9-mediated split-GFP tagging strategy (D. Kamiyama et al., 2016; R. Kamiyama et al., 2021; Kondo et al., 2020), we profiled the spatial distribution of endogenous Dop1R1 and Dop2R proteins in KCs, the PAM, and the PPL1 DANs.

Results

Co-expression of Dop1R1 and Dop2R genes in the adult Drosophila brain

To compare the expression of different dopamine receptor genes in detail, we used fluorescent reporter knock-ins of the endogenous Dop1R1 and Dop2R genes (Kondo et al., 2020). Both lines labelled many neuropils including the MB (Figure 1A and 1B), and the overlapping expression of these receptors is consistent with our previous quantification of GAL4-positive cells for both genes (58,049 and 68,528 for Dop1R1 and Dop2R, respectively, out of 118,331 brain cells; Kondo et al., 2020). Double labelling of Dop1R1 and Dop2R with T2A-LexA and T2A-GAL4, respectively, revealed cells with overlapping and differential expression (Fig. 1D-G; see also Kondo et al., 2020). The PAM cluster of DANs expressed both Dop1R1 and Dop2R (Figure 1D). On the other hand, most of the DANs in the PPL1 cluster strongly expressed Dop2R, but Dop1R1 only weakly (Figure 1E). We further found that Dop1R1, but not Dop2R, was highly expressed in the ring neurons of the ellipsoid body (Figure 1F; Hanesch et al., 1989) and the neuropil ensheathing glia (Figure 1G; Awasaki et al., 2008). In conclusion, Dop1R1 and Dop2R genes are co-expressed in the PAM neurons, KCs and a subset of the PPL1 neurons.

Co-expression of Dop1R1 and Dop2R genes in adult Drosophila brain.

(A and B) The expression of Dop1R1-T2A-GAL4 and Dop2R-T2A-GAL4 visualized by UAS-mCD8::GFP (green) The neuropil was stained by using Brp::SNAP (red). Maximum-intensity projections of the whole brain.

(C) Schematic of the KCs and the MB-innervating dopamine neurons from the PAM and PPL1 clusters.

(D-G) Double labelling of Dop1R1 and Dop2R gene expression visualized by Dop1R1-T2A-LexA/lexAop-rCD2::GFP (green) and Dop2R-T2A-GAL4/UAS-CD4::tdTomato (red). Dopamine neurons were immunostained with anti-TH antibody (blue). Single optical sections. Cell bodies of the PAM cluster (D) and the PPL1 cluster (E). The ellipsoid body (F) and ensheathing glia (G). Scale bars, 50 µm (A and B), 5 µm (D-G).

Quantification of subcellular enrichment of endogenous proteins in target cells

The overlapping expression of Dop1R1 and Dop2R genes prompted us to examine the subcellular localization of these receptor proteins. To elucidate the localization of these broadly expressed receptors (Figure 1), we took advantage of split-GFP tagging of endogenous proteins (D. Kamiyama et al., 2016; Kondo et al., 2020). By adding GFP11 tags to the C-termini of the Dop1R1 and Dop2R proteins, their intracellular distribution can be visualized specifically in cells expressing GFP1-10 through split-GFP reconstitution (Figure 2A).

Cell type-specific visualization of endogenous proteins with GFP11 tag.

(A) Principle of cell type specific fluorescent labelling of target proteins by GFP11 tag. GFP1-10 and membrane marker CD4::tdTomato are expressed in the target cells by GAL4/UAS system. In the target cells, reconstitution of GFP occurs on the endogenous proteins tagged with GFP11.

(B) As an example, DopEcR::GFP11 is visualized in KCs using MB-GeneSwitch, a ligand-inducible GAL4 driver. A merged image of reconstituted GFP (green) and cellular membrane visualized by UAS-CD4::tdTomato (magenta). Maximum intensity projection of the whole left MB.

(C) The workflow for visualizing subcellular protein enrichment by localization index (LI). A single sagittal section of the MB calyx and peduncle is shown. The ratio of reconstituted GFP to membrane signal in the left image is calculated and normalized by the mean intensity of all voxels to provide LI. In the middle image, LI is color-coded so that red represents local receptor enrichment. In the right image, the LI color is mapped on the membrane signal.

To quantify the subcellular enrichment of the receptors, we devised Localization Index (LI). Briefly, LI is the normalized ratio of the reconstituted GFP (rGFP) signal to the reference membrane marker signal (CD4::tdTomato). If the target and reference proteins had the identical distribution, LI would be 1 everywhere. More or less concentrations of rGFP result in the LI larger or smaller than 1, respectively (Figure 2; see also Methods for detail). LI was visualized as the color of the membrane signals (Figure 2C). As proof of principle, mapping the LI of DopEcR-rGFP signals in KCs highlighted enrichment in the proximal peduncle (Figure 2B, C), which is consistent with the previous report (Kondo et al., 2020). This representation thus visualizes the subcellular enrichment of the targeted receptors in the plasma membrane of GAL4-expressing cells.

Partial colocalization of Dop1R1 and Dop2R proteins

First, we compared the localization of Dop1R1 and Dop2R proteins in KCs, where both receptor genes were highly expressed (Figure 1D). Both receptors were distributed throughout KC membranes with predominant enrichment in the MB lobes (Figure 3B-E). This lobe enrichment was more pronounced for Dop2R than Dop1R1 (Figure 3F and 3G). LI measurements indicated that Dop2R was approximately twice as enriched in the lobe as the average density across the whole neuron, while the lobe enrichment of Dop1R1 was about 1.5 times the average (Figure 3H).

Subcellular localization of Dop1R1 and Dop2R in the Kenyon Cells.

Subcellular localization of Dop1R1 and Dop2R in the KCs is visualized by GFP11-tag. MB-GeneSwitch was used to express GFP1-10 and CD4::tdTomato in the KCs. Receptors in αβ KCs were visualized after 12 hours of RU486 feeding (see Materials and Methods).

(A) Schematic of the distribution of presynapses and postsynaposes in the KCs.

(B-G) Enrichment of Dop1R1 and Dop2R in the MB lobe. Frontal views of the MB (B and D) and sagittal views of the MB (C, E, F and G) of the maximum-intensity projections of the whole left MB are shown. Reconstituted GFP signals for both Dop1R1:: and Dop2R::GFP11 distributed throughout the MB lobe and the calyx (B-E). Visualization by LI showed more pronounced enrichment of Dop2R (G) than Dop1R1 (F) in the lobe as exemplified in the tip (arrows, 10 µm square, magnified in the insertion). Scalebars, 20 µm (B-G).

(H) Mean LI for Dop1R1 and Dop2R in the calyx, the peduncle, the tip of α and β lobes. Student’s t test was performed to compare LI of Dop1R1 and Dop2R in each region (N = 3). Error bars; SEM. * p< 0.05.

As KCs have major presynaptic sites in the MB lobes, we examined receptor localization to the active zones (AZ) using antibodies against Brp, an AZ scaffolding protein (Wagh et al., 2006). Interestingly, Dop1R1 and Dop2R proteins in KCs were tightly associated with the Brp puncta in the lobe (Figure 4A and 4B). The distribution of Dop2R, compared to Dop1R1, showed a greater association, as measured by the overlap of rGFP and Brp signals (Figure 4C). This was consistent with the higher lobe enrichment of Dop2R (Figure 3H). The AZ localization (Christiansen et al., 2011) of Dop1R1 and Dop2R was similar in the calyx (Figure 4 – figure supplement 1), although the identity of presynaptic neurons is less clear.

Presynaptic localization of Dop1R1 and Dop2R in Kenyon Cells and giant neurons.

(A and B) Double labelling of dopamine receptors (green) and Brp (magenta). Single focal slice at the tip of the vertical lobe. Insertions are the magnified images of white squares in the main panel.

(C) Colocalization analysis showed that Dop2R had significantly higher colocalization with Brp than Dop1R1 (p = 0.0259, Mann-Whitney U test, N = 6). Error bars; SEM.

(D-E) Punctate Brp expression in a giant neuron culture differentiated from cytokinesis-arrested neuroblasts of OK371-GAL4/UAS-mCD8::GFP embryos. Aggregated Brp condensates (magenta) were observed in the neurite terminals of the cells marked with mCD8::GFP (cyan) in (E).

(F-I) Double labelling of dopamine receptors (green) and the AZs (magenta). Dop1R1::Venus

(F and G) or Dop2R::Venus (H and I) was crossed with Brp::SNAP.

Scale bars, 20 µm (A and B), 10 μm (D, F and H), 5 μm (E), 1 μm (G and I).

Dop2R and Dop1R1 at the presynaptic sites of Kenyon Cells in the calyx.

(A-F) Double labelling of dopamine receptors (green) and Brp (magenta). From the single focal slice of the calyx (A and D), white squares including dendritic claws (B and E) and proximal dendrites (E and F) are magnified.

Scalebars, 20 µm (A and D).

To better resolve the presynaptic localization of DopR1 and Dop2R, we turned to “giant” Drosophila neurons differentiated from cytokinesis-arrested neuroblasts in culture (Wu et al., 1990). These “giant” neurons exhibit action potential firing patterns and neurotransmission similar to those described in mature neurons (Saito & Wu, 1991; Yao and Wu, 2001; Yao et al., 2000). Due to the large size of the cells, this system has advantages for investigating the microanatomy of neurons (Saito & Wu, 1991; Wu et al., 1990; Yao et al., 2000). In the giant neurons from the Brp::SNAP embryos (Kohl et al., 2014), we found punctate aggregation of Brp in the terminals of neurites, indicating the proper assembly of the AZ cytomatrix (Figure 4D and 4E). We confirmed that the giant neurons express Venus-tagged Dop1R1 and Dop2R (Figure 4F-4I). Strikingly, double labelling of these receptors with Brp::SNAP showed the association, but not overlap, with the Brp puncta, suggesting perisynaptic receptor accumulation (Figure 4F-4I). A closer investigation revealed a tighter association between Dop2R and Brp (Figure 4G and 4I). Presynaptic enrichment of these receptors in the giant neurons was strikingly similar to their localization in KCs, further substantiating presynaptic dopaminergic modulation.

D2-like receptors in mammals are expressed in DANs and act as autoreceptors, which mediates feedback signals by receiving neurotransmitters released from the neuron on which the receptors reside (Ford, 2014). In Drosophila, multiple dopamine receptor genes are expressed in DANs (Figure 1; Deng et al., 2019; Aso et al., 2019; Kondo et al., 2020), but it is unclear if these receptors function as autoreceptors. We therefore examined the subcellular localization of Dop1R1 and Dop2R in the PAM cluster DANs with a particular focus on their presynaptic terminals. The PAM neurons are polarized; presynaptic proteins are abundantly enriched in the MB lobe and barely detected in dendrites (Figure 5A-5C). We visualized Dop1R1 and Dop2R proteins in the PAM cluster DANs using R58E02-GAL4, and both were localized to the terminals in the MB lobes (Figure 5A and 5B). Representation of LI for Dop1R1 and Dop2R in the PAM neurons again showed stronger presynaptic enrichment of Dop2R than Dop1R1 (Figure 5F and 5G). Dop2R was strongly enriched at β’1 compartment showing significantly higher LI than Dop1R1 (Figure 5H-5J). Higher magnification revealed the accumulation of both Dop1R1 and Dop2R in the boutons (Figure 5K and 5L). The presynaptic localization of Dop1R1 and Dop2R in DANs suggests that both receptors mediate feedback regulation.

Subcellular localization of Dop1R1 and Dop2R in dopamine neurons.

(A-B) Maximum-intensity projection image showing the distribution of presynaptic sites in the PAM neurons. (A) R58E02-GAL4 was used to express UAS-mCD8::GFP (magenta) and UAS-nSyb-CLIP (magenta). (B) Visualization by LI showing enrichment of nSyb signals in the lobe projection of the PAM neurons.

(C) Illustrated projection pattern of the PAM neurons. Red puncta indicate the sparse distribution of presynaptic sites in dendrites.

(D-L) Subcellular localization of GFP11-tagged Dop1R1 and Dop2R in the PAM neurons. R58E02-GAL4 was used to express UAS-GFP1-10and UAS-CD4::tdTomato in the PAM neurons.

(D and E) Reconstituted GFP signals of Dop1R1::GFP11 (D) and Dop2R::GFP11 (E) in the PAM neurons. Maximum-intensity projections of the left hemisphere including the MB lobe, SMP, SIP and SLP (superior medial, intermediate, and lateral protocerebrum, respectively). (F and G) Stronger presynaptic enrichment of Dop2R (G) than that of Dop1R1 (F) in the PAM dopamine neurons visualized by LI. β’1 compartment is magnified in the insertion image.

(H-J) R15A04-GAL4 was used to measure LI in the PAM-β’1 neuron. (H and I) Presynaptic terminals of the PAM-β’1 neurons are shown (dashed line). (H) Mean LI for Dop1R1 and Dop2R in β’1 (p = 0.0056, Mann-Whitney U test, N = 9). Error bars; SEM.

(K and L) A single optical slice of γ5 compartment in the MB lobe obtained using Airyscan. Merged image of reconstituted GFP (green) and CD4::tdTomato (magenta). Insertions are the magnified images of the presynaptic boutons of PAM-γ5 (white squares).

Scale bars, 20 µm (A-B, D-I), 5 µm (K and L).

State-dependent and bidirectional modulation of dopamine receptor expression in the PAM and PPL1 DANs

The activity of MB-projecting DANs is reported to be dynamic and sensitive to feeding states (Ichinose et al., 2017; Liu et al., 2012; Plaçais & Preat, 2013; Senapati et al., 2019; Siju et al., 2020; Tsao et al., 2018; Yamagata et al., 2016). We therefore examined if starvation alters the protein expression of Dop1R1 and Dop2R in these MB-projecting DANs (Figure 6). We found a significant elevation of Dop1R1 in the terminals of PAM-γ5 upon starvation for 10 hours or longer (Figure 6E). In contrast, starvation did not increase Dop2R in PAM-γ5 but rather tended to decrease, if at all (Figure 6E). We found similar starvation-dependent changes in Dop1R1 and Dop2R levels in other PAM neurons (Figure6 – figure supplement 1A-H). These results together suggest that starvation enhances presynaptic dopamine signaling in the reward-related PAM neurons by shifting the balance of Dop1R1 and Dop2R.

Bidirectional modification of dopamine receptor expression in dopamine neurons.

(A) Schematic diagram of the starvation protocol. (B) Schematic illustration of the MB projection of the PAM and PPL1 dopamine neurons marked by R15A04-GAL4 and TH-GAL4. (C and D) Dop1R (C) and Dop2R (D) in the presynaptic terminals of PAM-γ5 before and after 48 hours of starvation.

(E) Quantification of dopamine receptor levels in the presynaptic terminals of PAM-γ5 after 0, 10, 24 and 48 hours of starvation (n = 6-13).

(F and G) Reconstituted GFP signals of Dop1R1::GFP11 (F) and Dop2R::GFP11 (G) in the PPL1 neurons. In the MB projections of the PPL1 neurons, Dop1R1 is expressed in only the α3 compartment (F). Dop2R is expressed in all MB projections (G). Maximum-intensity projections of the MB lobe.

(H and I) Dop1R (H) and Dop2R (I) in the presynaptic terminals of PPL1-α3 before and after 24 hours of starvation.

(J) Quantification of the dopamine receptor levels in the presynaptic terminals of PPL1-α3 after 0, 10 and 24 hours of starvation (n = 7-10).

Interaction effects between genotypes and starvation time on protein levels were tested by Two-way ANOVA (E and J).

Scale bar, 10 µm (C, D, H and I), 20 µm (F and G).

Error bars; SEM (E and J). * p< 0.05, ** p< 0.01, *** p< 0.001, n.s. = not significant.

Starvation-dependent change of dopamine receptors in PAM and PPL1.

(A-F) Dop1R (A-C) and Dop2R (D-F) in the presynaptic terminals of PAM-α1 (A and D), PAM-β2 (B and E) and PAM-β’1 (C and F) before and after 48 hours of starvation.

(G and H) Quantification of dopamine receptor levels in the presynaptic terminals of the PAM neurons after 0, 10, 24 and 48 hours of starvation (n = 6-13).

(I-K) Dop2R in the presynaptic terminals of PPL1-α’2 (I), PPL1-γ1pedc (J) and PPL1-γ2 (K) before and after 24 hours of starvation.

(L) Quantification of Dop2R levels in the presynaptic terminals of the PPL1 neurons after 0, 10 and 24 hours of starvation (n = 7-10).

Scale bar, 10 µm (A-F, I-K). Error bars; SEM (G, H and L). * p< 0.05, ** p< 0.01, *** p< 0.001, n.s. not significant.

DANs of the PAM and PPL1 clusters exert distinct, largely opposite behavioral functions (Claridge-Chang et al., 2009; Liu et al., 2012). Therefore, plasticity in the PPL1 neurons may differ from that in the PAM neurons. To test this hypothesis, we examined the starvation-dependent change in Dop1R1 and Dop2R protein expression in the PPL1 neurons. To this end, we visualized GFP11-tagged receptors by expressing GFP1-10 using TH-GAL4 (Friggi-Grelin et al., 2003). We detected Dop2R proteins in all MB projections of the PPL1 neurons, whereas Dop1R1 proteins were only detectable in the terminals of the PPL1-α3 neuron (Figure 6F and 6G). Strikingly, the starvation-induced changes in the PPL1 neurons were opposite to those in the PAM: Dop2R, but not Dop1R1, significantly increased in the α3 compartment (Figure 6J), and the same tendency was observed in the other PPL1 neurons (Figure6 – figure supplement 1I-L). Taken together, starvation induces opposite responses of dopamine receptor expression in the PPL1 and PAM DANs, and we propose that these changes shift the balance of dopamine output from these subsets (Figure 7A).

The dual dopaminergic feedback regulating starved-state dependent expression of appetitive behavior.

(A) A working model showing the role of the dual dopaminergic feedback regulation. In starved state, increased Dop1R1 in PAM neurons and increased Dop2R in PPL1 neurons changes the balance between the synaptic outputs from these DANs to favor appetitive behavior.

(B) According to the model, loss of Dop2R in PPL1 upregulates output from PPL1 to attenuate appetitive behavior in starved flies.

(C) Knockdown of Dop2R in the PPL1 neurons by MB504B-GAL4 reduced 3-hour appetitive memory performance (n=14-15). Error bars; SEM. * p< 0.05.

The starvation-induced increase of Dop2R in the PPL1 terminals suggests enhanced negative feedback (Krashes et al., 2009). If so, loss of Dop2R would impair appetitive memory in starved flies, since inhibition of PPL1 output is necessary for appetitive memory expression (Krashes et al., 2009; Pavlowsky et al., 2018) (Figure 7B). To test this hypothesis, we examined appetitive olfactory associative memory by transgenic knockdown of Dop2R specifically in the PPL1 cluster neurons using MB504B-GAL4 (Vogt et al., 2014). Indeed, appetitive memory was impaired (Figure 7C), demonstrating the significance of the negative feedback regulation of the PPL1 in appetitive behavior expression. These results are in line with the state-dependent changes in the physiology of these DANs (Siju et al., 2020; Tsao et al., 2018). Taken together, the opposite presynaptic modulation of the PPL1 and PAM neurons through the antagonistic dopamine autoreceptors determines the balance of the MB output to control appetitive behaviors.

Discussion

The present study demonstrated the expression, subcellular localization, and dynamics of endogenous dopamine receptors, specifically Dop1R1 and Dop2R, in the MB circuit of the fly brain. Previously, intronic insertions of fluorescent tags have been used to label endogenous proteins in a cell-type specific manner (Fendl et al., 2020; R. Kamiyama et al., 2021; Williams et al., 2019). CRISPR/Cas9-based insertion is advantageous in terms of flexibility in fusion sites compared to intronic insertions (Fendl et al., 2020; R. Kamiyama et al., 2021; Williams et al., 2019). This enables the C-terminal tagging of dopamine receptors that do not perturb protein localization and function (Kondo et al., 2020).

The split-GFP reconstitution strategy enables sensitive detection of low-abundant endogenous proteins by the tandem multimerization of the GFP11 tags, as shown in other epitope tags such as SunTag and smFPs (Tanenbaum et al., 2014; Viswanathan et al., 2015). The background fluorescence of the split-GFP fragments was practically negligible. For example, Dop1R1-rGFP signal was observed in fewer cells than GFP1-10-expressing PPL1 neurons as expected from the gene expression pattern (D. Kamiyama et al., 2016; Kondo et al., 2020) (Figure 6). This approach can thus be applicable to monitor localization and dynamics of endogenous proteins of low abundance.

To quantify subcellular receptor localization, we here devised LI, which measures the relative enrichment of tagged proteins over the membrane markers. Quantitative localization analysis with LI revealed fine distinctions in presynaptic enrichment between Dop1R1 and Dop2R (figure 3 and 5). Dop2R showed a stronger association with AZ, while Dop1R1 was distributed more broadly along axons and terminals (Figure 4). Similar but distinct subcellular enrichment of Dop1R1 and Dop2R therefore suggests joint action in intricate regulation of presynaptic dopamine signaling.

Spatial regulation of dopamine signaling through distinct receptor subtypes

How do Dop1R1 and Dop2R function in presynaptic terminals? Differential AZ enrichment suggests differential contributions to the spatial regulation of dopamine response. Dopamine release is not confined to the synaptic cleft, but involves extracellular diffusion (Rice & Cragg, 2008) and extrasynaptic release (Liu & Kaeser, 2019). Such volume transmission can increase the number of target synapses by 10 times (Li et al., 2020; Takemura et al., 2017). The broader distribution of Dop1R1 thus enables to respond to the trace amount of extrasynaptic dopamine. On the other hand, the tight association of Dop2R with the AZ underscores the importance of reducing the noise of second messenger signaling likely through inhibiting activities of adenylyl cyclase and voltage-gated calcium channels (Beaulieu & Gainetdinov, 2011; Hearn et al., 2002). These second messengers, cAMP and Ca2+, were reported to regulate distinct aspects of presynaptic functions including release probability and vesicle recycling (Ehmann et al., 2018; Kuromi & Kidokoro, 2000, 2005). The spatial configuration of Dop1R1 and Dop2R may thus contribute to the sensitivity and precision in dopaminergic modulation of the synaptic vesicle cycle.

The dual autoreceptor system may shape dopamine release

Presynaptic localization of Dop1R1 and Dop2R in DANs strongly suggests their functions as autoreceptors. Reported functions of inhibitory dopamine autoreceptors range from negative regulations of the amplitude and kinetics of dopamine transmission, decreasing dopamine synthesis, to altering the uptake of released dopamine (Ford, 2014; Shin & Venton, 2022; Vickrey & Venton, 2011). Considering tight presynaptic Dop2R enrichment in the PAM and PPL1 DANs, such negative feedback should prevent overactivation of second messenger signaling at the AZ, where the ligand concentrations should be highest. In contrast, the presence of Dop1R1 in DAN terminals was unexpected, adding another layer of presynaptic regulations in dopamine release. Although it must be functionally verified if presynaptic Dop1R1 acts as autoreceptors in the PAM and PPL1 DANs, it should provide feedback for the dopamine release. This positive feedback would be particularly necessary in the regions where extracellular dopamine concentrations are low. Indeed, presynaptic Dop1R1 was undetectable in the PPL1 DNAs, which were reported to have high spontaneous activities (Feng et al., 2021; Plaçais & Preat, 2013). Furthermore, we noticed that compositions of presynaptic Dop1R1 and Dop2R vary among individual DANs (Figure 5 and 6). This dual autoreceptor system may thus fine-tune the amplitude and kinetics of dopamine release depending on the cellular function.

Dopamine receptor expression is reported to be associated with prolonged exposure to psychoactive substances, such as caffeine and ethanol (Andretic et al., 2008; Kanno et al., 2021; Kondo et al., 2020; Petruccelli et al., 2018). This study further showed starvation-dependent changes of Dop1R1 and Dop2R in DAN terminals (Figure 7). Strikingly, starvation responses of presynaptic Dop1R1 and Dop2R were differential depending on the DAN cell types (Figure 7). We therefore propose the state-dependent bidirectional modulation of the MB output through the dual dopamine autoreceptor system in the PAM and PPL1 neurons (Figure 7A). Starvation changes the dual autoreceptor system to put more weight on the PAM output over PPL1 to control the expression of appetitive behavior (Figure 7A). This shifted balance explains the state-dependent changes in the presynaptic activity of these two clusters of DANs and recruited MBONs for appetitive behavior (Aso, Sitaraman, et al., 2014; Ichinose et al., 2021; Landayan et al., 2018; Owald et al., 2015; Siju et al., 2020; Tsao et al., 2018).

Materials and Methods

Resource tables

Flies

Flies were raised on standard cornmeal food at 25°C under a 12:12 h light-dark cycle (ZT0 at 8 AM) for all experiments. The GAL4-UAS system was used to express the transgenes of interest in specific neuron subtypes. Flies carrying GAL4 were crossed to another strain carrying UAS reporters, and F1 progenies were used in the experiments. To visualize GFP11-tagged dopamine receptors in the specific cell-types, female fly strains carrying UAS-CD4::tdTomato, UAS-GFP1- 10 and GAL4 driver was crossed to male fly strains carrying Dop1R1::GFP11or Dop2R::GFP11 and F1 progenies were used. To make the giant neuron culture, embryos from Brp::SNAP or the F1 progeny of OK371-GAL4 (BDSC #26160) crossed with Brp::SNAP, UAS-mCD8::GFP, UAS-nSyb-CLIP flies were used. For the fly lines used in our manuscript, see the Resource Tables.

Brain dissection and Immunohistochemistry

Male flies with the specific genotype were sorted under CO2 anesthesia and kept in a food vial for recovery at least 24 h prior to the dissection. Fly brains were dissected 3-7 days after eclosion in ice-cold Phosphate-buffered saline (PBS). After dissection, brains were kept in ice-cold PBS with 2% paraformaldehyde (PFA) for up to 30 min. For fixation, brains were incubated in 2 % PFA in PBS for 1 hour at room temperature. Fixed brains were washed in PBS with 0.1% Triton X-100 (PBT) for 10 min 3 times. For rGFP imaging, fixed brains were mounted in SeeDB2 (Ke et al., 2016), and native fluorescence was imaged without signal amplification.

For chemical tagging reaction (Figure 5A and 5B), brains were incubated in PBT (0.3 % Triton X-100 in PBS) containing CLIP-Surface 547 substrate (1µM, S9233S; New England Biolabs Inc.) for 15 min at room temperature. Brains were washed in PBT for 10 min 3 times.

For immunohistochemistry, fixed brains were blocked in 3 % normal goat serum (NGS) for 30 min at room temperature. Brains were incubated in primary and secondary antibodies diluted in PBT with 1 % NGS over two nights at 4 ℃, respectively. After each step, brains were washed three times in PBT for longer than 20 min at room temperature and mounted in 86% glycerol.

Following primary antibodies were used: mouse anti-TH (1:100; 22941; ImmunoStar Inc., Hudson, WI, USA), mouse anti-Brp (1:40; nc82; Developmental Studies Hybridoma Bank, Univ. Iowa). Secondary antibodies: Alexa Fluor 633 goat anti-mouse (1:400 for anti-TH, 1:200 for nc82; A-21052; Invitrogen, Waltham, MA, USA).

Fluorescent imaging

For image acquisition, whole-mount brains were scanned with the Olympus FV1200 confocal microscope with following objective lens; 20x oil (NA = 0.85, UPLSAPO20XO, Olympus), 30x silicone (NA = 1.05, UPLSAPO30XS, Olympus), 40x oil (NA = 1.3, UPLFLN40XO, Olympus), or 60x oil (NA = 1.42, PLAPON60XO, Olympus). Z-stack images were acquired with following voxel sizes; 0.468 x 0.468 x 0.84 µm3 for 20x oil (Figure 1A and 1B), 0.099 x 0.099 x 0.44 μm3 for 30x silicone (Figure 1D-1H), 0.194 x 0.194 x 0.52 μm3 for 40x oil (Figure 3B-3G, 6C-6D, 6F-6I and Figure 6 – figure supplement 1 A-F, I-K), and 0.101 x 0.101 x 0.45 μm3 for 60x oil (Figure 4A, 4B and Figure 4 – figure supplement 1 A-F).

For high resolution imaging in Figure 5I-J, Airyscan imaging was performed on Zeiss LSM800 with 63x oil objective lens (NA = 1.40, Plan-Apochromat). A single focal slice was scanned with a pixel size of 0.0425 x 0.0425 μm2.

Image analysis

All image analysis were conducted on Fiji (Schindelin et al., 2012).

To visualize the subcellular localization of dopamine receptors in defined neurons, we devised the localization index (LI). Relative receptor density was calculated in each voxel by dividing the receptor (reconstituted GFP) signals by the corresponding membrane signals (CD4::tdTomato). To set ROI, any voxels devoid of membrane signal were censored. The local density was normalized by the mean values in the ROI. For the visual representation, the reference membrane signals were colored according to the normalized LI (Figure 2C, middle). To quantify mean LI in each subcellular region (Figure 3H), ROI was manually set at single focal slice that include the largest area of interested region.

Three-dimensional iterative deconvolution with the Richardson-Lucy algorithm was applied to some stack images (Figure 4A-B, Figure 4 – figure supplement 1). To generate the theoretical PSF for deconvolution, the ImageJ plug-in diffraction PSF 3D was used. Deconvolved images were used for following colocalization analysis.

To analyze the colocalization of dopamine receptors and Brp (Figure 4C), both channels were binarized with the auto-thresholding on Fiji. The ratio of overlap between rGFP and Brp signals over rGFP signals was calculated. Square ROI (10.1 x 10.1 µm, 100 x 100 pixel) at single focal slice was sampled within the tip of the vertical lobes for colocalization analysis.

Embryonic giant neuron culture

Multinucleated giant neurons from neuroblasts were generated as described previously (Wu et al., 1990). Briefly, the interior of a gastrula from stage 6/7 embryo was extracted with a glass micropipette and dispersed into a drop of culture medium (∼ 40 μm) on an uncoated coverslip. The culture medium contained 80% Drosophila Schneider’s insect medium (S0146-100ML, Merck KGaA, Darmstadt, Germany) and 20% fetal bovine serum (F2442-100ML, Merck KGaA, Darmstadt, Germany), with the addition of 50 μg/ml streptomycin, 50 U/ml penicillin (all from Sigma, St. Louis, MO, USA) and cytochalasin B (CCB; 2 μg/ml; Sigma, St. Louis, MO, USA). CCB was removed by washing with CCB-free medium 1 d after plating. All cultures were grown in a humidified chamber.

To label Brp::SNAP, cells were incubated in fluorescent SNAP substrate diluted in the culture medium (1:5000, SNAP-Cell 647-SiR, S9102S, Ipswich, MA, USA) for half an hour at RT. For double labeling with nSyb-CLIP, cells were first incubated in fluorescent CLIP substrate diluted in the culture medium (1:5000, CLIP-Cell™ TMR-Star, S9219S, Ipswich, MA, USA) for half an hour at RT. Subsequently, the same cells were labeled by the SNAP substrate to minimize the crosstalk. After the incubation, cells were washed with the culture medium and subjected to confocal scanning with Leica TCS SP8 microscopy equipped with a 40x oil immersion objective (HC PL APO 40x/1.30 Oil PH3 CS2, Leica). Acquired images were then processed with the LIGHTNING software.

Behavioral assays

The experimental protocols were as described previously (Ichinose and Tanimoto., 2016). For appetitive conditioning, a group of approximately 50 flies in a training tube alternately received 3-octanol (3OCT; Merck) and 4-methylcyclohexanol (4MCH; Sigma-Aldrich), for 1 minute in a constant air flow with or without reward with an interval of 1 minute between the two odor presentations. These odorants were diluted to 1.2% and 2% with paraffin wax (Sigma-Aldrich), respectively. Dried 2 M sucrose (Sigma-Aldrich) on a piece of filter paper was used as the reward. Flies were starved in the presence of wet tissue paper for 24 hours before appetitive conditioning. For testing, flies were given a choice between CS+ and CS-, which distribution in the plexiglass T-maze (Ichinose and Tanimoto, 2016) was video-recorded one frame per second for 2 minutes with the CMOS cameras (GS3-U3-51S5M, FLIR) with infrared LED illumination (Ichinose & Tanimoto, 2016). Flies were forced to stay on the floor by applying Fluon (Insect-a-Slip PTFE30, BioQuip Products) on the side and top of T-maze arms.

Learning index was then calculated by counting the number of flies in each arm using ImageJ macro as described previously (Ichinose and Tanimoto, 2016) with the following formula (Tempel et al., 1983): where #cs+ and #cs− imply the number of flies in the CS+ and the CS-arms, respectively. Learning index was calculated for every second and was averaged for the last 60 s of the 120 s test session. An average of a pair of reciprocally trained groups was used as a single data point.

Statistics

For multiple comparison (Figure 6 and Figure 6 – figure supplement 1), statistics were performed by Prism5 (GraphPad). Data were always first tested for normality (Shapiro-Wilk test) and for homoscedasticity (Spearman’s test or Bartlett’s test). If these assumptions were not violated, parametric tests (one-way ANOVA, followed by Dunnett’s post hoc pairwise test) were applied. If data did not suffice the assumptions for parametric tests, nonparametric tests (Kruskal-Wallis, followed by Dunn’s post hoc pairwise test) were performed. For comparison of the two groups, Student’s t test (Figure 3H) or Mann-Whitney U test (Figure 4C, 5H) was performed on R Statistical Software.

Bars and error bars represent means and SEM, respectively, in all figures. For all figures, significance corresponds to the following symbols: * p < 0.05; ** p < 0.01; *** p < 0.001; n.s. not significant.

eLife Assessment

This study uses state-of-the-art methods to label endogenous dopamine receptors in a subset of Drosophila mushroom body neuronal types. The authors report that DopR1 and Dop2R receptors, which have opposing effects in intracellular cAMP, are present in axons termini of Kenyon cells, as well as those of two classes of dopaminergic neurons that innervate the mushroom body indicative of autocrine modulation by dopaminergic neurons. Additional experiments showing opposing effects of starvation on DopR1 and DopR2 levels in mushroom body neurons are consistent with a role for dopamine receptor levels increasing the efficiency of learned food-odour associations in starved flies. Supported by solid data, this is a valuable contribution to the field.

Reviewer #1 (Public Review):

Summary:

This is an important and interesting study that uses the split-GFP approach. Localization of receptors and correlating them to function is important in understanding the circuit basis of behavior.

Strengths:

The split-GFP approach allows visualization of subcellular enrichment of dopamine receptors in the plasma membrane of GAL4-expressing neurons allowing for a high level of specificity.

The authors resolve the presynaptic localization of DopR1 and Dop2R, in "giant" Drosophila neurons differentiated from cytokinesis-arrested neuroblasts in culture as it is not clear in the lobes and calyx.

Starvation-induced opposite responses of dopamine receptor expression in the PPL1 and PAM DANs provide key insights into models of appetitive learning.

Starvation-induced increase in D2R allows for increased negative feedback that the authors test in D2R knockout flies where appetitive memory is diminished.

This dual autoreceptor system is an attractive model for how amplitude and kinetics of dopamine release can be fine-tuned and controlled depending on the cellular function and this paper presents a good methodology to do it and a good system where the dynamics of dopamine release can be tested at the level of behavior.

Weaknesses:

LI measurements of Kenyon cells and lobes indicate that Dop2R was approximately twice as enriched in the lobe as the average density across the whole neuron, while the lobe enrichment of Dop1R1 was about 1.5 times the average, are these levels consistent during different times of the day and the state of the animal. How were these conditions controlled and how sensitive are receptor expression to the time of day of dissection, staining, etc.

The authors assume without discussion as to why and how presynaptic enrichment of these receptors is similar in giant neurons and MB.

Figures 1-3 show the expensive expression of receptors in alpha and beta lobes while Figure 5 focusses on PAM and localization in γ and β' projections of PAM leading to the conclusion that pre-synaptic dopamine neurons express these and have feedback regulation. Consistency between lobes or discussion of these differences is important to consider.

Receptor expression in any learning-related MBONs is not discussed, and it would be intriguing as how receptors are organized in those cells. Given that these PAMs input to both KCs and MBONs these will have to work in some coordination.

Although authors use the D2R enhancement post starvation to show that knocking down receptors eliminated appetitive memory, the knocking out is affecting multiple neurons within this circuit including PAMs and KCs. How does that account for the observed effect? Are those not important for appetitive learning?

The evidence for fine-tuning is completely based on receptor expression and one behavioral outcome which could result from many possibilities. It is not clear if this fine-tuning and presynaptic feedback regulation-based dopamine release is a clear possibility. Alternate hypotheses and outcomes could be considered in the model as it is not completely substantiated by data at least as presented.

Reviewer #2 (Public Review):

Summary:

Hiramatsu et al. investigated how cognate neurotransmitter receptors with antagonizing downstream effects localize within neurons when co-expressed. They focus on mapping the localization of the dopaminergic Dop1R1 and Dop2R receptors, which correspond to the mammalian D1- and D2-like dopamine receptors, which have opposing effects on intracellular cAMP levels, in neurons of the Drosophila mushroom body (MB). To visualize specific receptors in single neuron types within the crowded MB neuropil, the authors use existing dopamine receptor alleles tagged with 7 copies of split GFP to target reconstitution of GFP tags only in the neurons of interest as a read-out of receptor localization. The authors show that both Dop1R1 and Dop2R, with differing degrees, are enriched in axonal compartments of both the Kenyon Cells cholinergic presynaptic inputs and in different dopamine neurons (DANs), which project axons to the MB. Co-localization studies of dopamine receptors with the presynaptic marker Brp suggest that Dop1R1 and, to a larger extent Dop2R, localize in the proximity of release sites. This localization pattern in DANs suggests that Dop1R1 and Dop2R work in dual-feedback regulation as autoreceptors. Finally, they provide evidence that the balance of Dop1R1 and Dop2R in the axons of two different DAN populations is differentially modulated by starvation and that this regulation plays a role in regulating appetitive behaviors.

Strengths:

The authors use reconstitution of GFP fluorescence of split GFP tags knocked into the endogenous locus at the C-terminus of the dopamine receptors as a readout of dopamine receptor localization. This elegant approach preserves the endogenous transcriptional and post-transcriptional regulation of the receptor, which is essential for studies of protein localization.

The study focuses on mapping the localization of dopamine receptors in neurons of the mushroom body. This is an excellent choice of system to address the question posed in this study, as the neurons are well-studied, and their connections are carefully reconstructed in the mushroom body connectome. Furthermore, the role of this circuit in different behaviors and associative memory permits the linking of patterns of receptor localization to circuit function and resulting behavior. Because of these features, the authors can provide evidence that two antagonizing dopamine receptors can act as autoreceptors within the axonal compartment of MB innervating DANs. The differential regulation of the balance of the two receptors under starvation in two distinct DAN innervations provides evidence of the role that regulation of this balance can play in circuit function and behavioral output.

Weaknesses:

The approach of using endogenously tagged alleles to study localization is a strength of this study, but the authors do not provide sufficient evidence that the insertion of 7 copies of split GFP to the C terminus of the dopamine receptors does not interfere with the endogenous localization pattern or function. Both sets of tagged alleles (1X Venus and 7X split GFP tagged) were previously reported (Kondo et al., 2020), but only the 1X Venus tagged alleles were further functionally validated in assays of olfactory appetitive memory. Despite the smaller size of the 7X split-GFP array tag knocked into the same location as the 1X venus tag, the reconstitution of 7 copies of GFP at the C terminus of the dopamine receptor, might substantially increase the molecular bulk at this site, potentially impeding the function of the receptor more significantly than the smaller, single Venus tag. The data presented by Kondo et al. 2020, is insufficient to conclude that the two alleles are equivalent.

The authors' conclusion that the receptors localize to presynaptic sites is weak. The analysis of the colocalization of the active zone marker Brp whole-brain staining with dopamine receptors labeled in specific neurons is insufficient to conclude that the receptors are localized at presynaptic sites. Given the highly crowded neuropil environment, the data cannot differentiate between the receptor localization postsynaptic to a dopamine release site or at a presynaptic site within the same neuron. The known distribution of presynaptic sites within the neurons analyzed in the study provides evidence that the receptors are enriched in axonal compartments, but co-labeling of presynaptic sites and receptors in the same neuron or super-resolution methods are needed to provide evidence of receptor localization at active zones. The data presented in Figures 5K-5L provides compelling evidence that the receptors localize to neuronal varicosities in DANs where the receptors could play a role as autoreceptors.

Given the highly crowded environment of the mushroom body neuropil, the analysis of dopamine receptor localization in Kenyon cells is not conclusive. The data is sufficient to conclude that the receptors are preferentially localizing to the axonal compartment of Kenyon cells, but co-localization with brain-wide Brp active zone immunostaining is not sufficient to determine if the receptor localizes juxtaposed to dopaminergic release sites, in proximity of release sites in Kenyon cells, or both.

Author response:

The following is the authors’ response to the original reviews.

eLife Assessment

This study uses state-of-the-art methods to label endogenous dopamine receptors in a subset of Drosophila mushroom body neuronal types. The authors report that DopR1 and Dop2R receptors, which have opposing effects in intracellular cAMP, are present in axons termini of Kenyon cells, as well as those of two classes of dopaminergic neurons that innervate the mushroom body indicative of autocrine modulation by dopaminergic neurons. Additional experiments showing opposing effects of starvation on DopR1 and DopR2 levels in mushroom body neurons are consistent with a role for dopamine receptor levels increasing the efficiency of learned food-odour associations in starved flies. Supported by solid data, this is a valuable contribution to the field.

We thank the editors for the assessment, but request to change “DopR2” to “Dop2R”. The dopamine receptors in Drosophila have confusing names, but what we characterized in this study are called Dop1R1 (according to the Flybase; aka DopR1, dDA1, Dumb) and Dop2R (ibid; aka Dd2R). DopR2 is the name of a different dopamine receptor.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

This is an important and interesting study that uses the split-GFP approach. Localization of receptors and correlating them to function is important in understanding the circuit basis of behavior.

Strengths:

The split-GFP approach allows visualization of subcellular enrichment of dopamine receptors in the plasma membrane of GAL4-expressing neurons allowing for a high level of specificity.

The authors resolve the presynaptic localization of DopR1 and Dop2R, in "giant" Drosophila neurons differentiated from cytokinesis-arrested neuroblasts in culture as it is not clear in the lobes and calyx.

Starvation-induced opposite responses of dopamine receptor expression in the PPL1 and PAM DANs provide key insights into models of appetitive learning.

Starvation-induced increase in D2R allows for increased negative feedback that the authors test in D2R knockout flies where appetitive memory is diminished.

This dual autoreceptor system is an attractive model for how amplitude and kinetics of dopamine release can be fine-tuned and controlled depending on the cellular function and this paper presents a good methodology to do it and a good system where the dynamics of dopamine release can be tested at the level of behavior.

Weaknesses:

LI measurements of Kenyon cells and lobes indicate that Dop2R was approximately twice as enriched in the lobe as the average density across the whole neuron, while the lobe enrichment of Dop1R1 was about 1.5 times the average, are these levels consistent during different times of the day and the state of the animal. How were these conditions controlled and how sensitive are receptor expression to the time of day of dissection, staining, etc.

To answer this question, we repeated the experiment in two replicates at different times of day and confirmed that the receptor localization was consistent (Figure 3 – figure supplement 1); LI measurements showed that Dop2R is enriched more in the lobe and less in the calyx compared to Dop1R1 (Figure 3D). The states of animals that could affect LI (e.g. feeding state and anesthesia for sorting, see methods) were kept constant.

The authors assume without discussion as to why and how presynaptic enrichment of these receptors is similar in giant neurons and MB.

In the revision, we added a short summary to recapitulate that the giant neurons exhibit many characteristics of mature neurons (Lines #152-156): "Importantly, these giant neurons exhibit characteristics of mature neurons, including firing patterns (Wu et al., 1990; Yao & Wu, 2001; Zhao & Wu, 1997) and acetylcholine release (Yao et al., 2000), both of which are regulated by cAMP and CaMKII signaling (Yao et al., 2000; Yao & Wu, 2001; Zhao & Wu, 1997)." In addition, we found punctate Brp accumulations localized to the axon terminals of the giant neurons (former Figure 4D and 4E). Therefore, the giant neuron serves as an excellent model to study the presynaptic localization of dopamine receptors in isolated large cells.

Figures 1-3 show the expensive expression of receptors in alpha and beta lobes while Figure 5 focusses on PAM and localization in γ and β' projections of PAM leading to the conclusion that presynaptic dopamine neurons express these and have feedback regulation. Consistency between lobes or discussion of these differences is important to consider.

In the revised manuscript, we show data in the γ KCs (Figure 4C, Figure 5 - figure supplement 1) in addition to α/β KCs, and demonstrate the consistent synaptic localization of Dop1R1 and Dop2R as in α/β KCs (Figure 4B and 5A).

Receptor expression in any learning-related MBONs is not discussed, and it would be intriguing as how receptors are organized in those cells. Given that these PAMs input to both KCs and MBONs these will have to work in some coordination.

The subcellular localization of dopamine receptors in MBONs indeed provides important insights into the site of dopaminergic signaling in these neurons (Takemura et al., 2017; Pavlowsky et al., 2018; Pribbenow et al., 2022). Therefore, we added new data for Dop1R1 and Dop2R in MBON-γ1pedc>αβ (Figure 6). Interestingly, these receptors are localized to in the dendritic projection in the γ1 compartment as well as presynaptic boutons (Figure 6).

Although authors use the D2R enhancement post starvation to show that knocking down receptors eliminated appetitive memory, the knocking out is affecting multiple neurons within this circuit including PAMs and KCs. How does that account for the observed effect? Are those not important for appetitive learning?

In the appetitive memory experiment (Figure 9C), we knocked down Dop2R only in the select neurons of the PPL1 cluster, and this manipulation does not directly affect Dop2R expression in PAMs and KCs.

Starvation-induced enhancement of Dop2R expression in the PPL1 neurons (Figure 8F) would attenuate their outputs and therefore disinhibit expression of appetitive memory in starved flies (Krashes et al., 2009). Consistently, Dop2R knock-down in PPL1 impaired appetitive memory in starved flies (Figure 9C). We revised the corresponding text to make this point clearer (Lines #224227).

The evidence for fine-tuning is completely based on receptor expression and one behavioral outcome which could result from many possibilities. It is not clear if this fine-tuning and presynaptic feedback regulation-based dopamine release is a clear possibility. Alternate hypotheses and outcomes could be considered in the model as it is not completely substantiated by data at least as presented.

The reviewer’s concern is valid, and the presynaptic dopamine tuning by autoreceptors may need more experimental support. We therefore additionally discussed another possibility (Lines #289-291): “Alternatively, these presynaptic receptors could potentially receive extrasynaptic dopamine released from other DANs. Therefore, the autoreceptor functions need to be experimentally clarified by manipulating the receptor expression in DANs.”

Reviewer #2 (Public Review):

Summary:

Hiramatsu et al. investigated how cognate neurotransmitter receptors with antagonizing downstream effects localize within neurons when co-expressed. They focus on mapping the localization of the dopaminergic Dop1R1 and Dop2R receptors, which correspond to the mammalian D1- and D2-like dopamine receptors, which have opposing effects on intracellular cAMP levels, in neurons of the Drosophila mushroom body (MB). To visualize specific receptors in single neuron types within the crowded MB neuropil, the authors use existing dopamine receptor alleles tagged with 7 copies of split GFP to target reconstitution of GFP tags only in the neurons of interest as a read-out of receptor localization. The authors show that both Dop1R1 and Dop2R, with differing degrees, are enriched in axonal compartments of both the Kenyon Cells cholinergic presynaptic inputs and in different dopamine neurons (DANs), which project axons to the MB. Co-localization studies of dopamine receptors with the presynaptic marker Brp suggest that Dop1R1 and, to a larger extent Dop2R, localize in the proximity of release sites. This localization pattern in DANs suggests that Dop1R1 and Dop2R work in dual-feedback regulation as autoreceptors. Finally, they provide evidence that the balance of Dop1R1 and Dop2R in the axons of two different DAN populations is differentially modulated by starvation and that this regulation plays a role in regulating appetitive behaviors.

Strengths:

The authors use reconstitution of GFP fluorescence of split GFP tags knocked into the endogenous locus at the C-terminus of the dopamine receptors as a readout of dopamine receptor localization. This elegant approach preserves the endogenous transcriptional and post-transcriptional regulation of the receptor, which is essential for studies of protein localization.

The study focuses on mapping the localization of dopamine receptors in neurons of the mushroom body. This is an excellent choice of system to address the question posed in this study, as the neurons are well-studied, and their connections are carefully reconstructed in the mushroom body connectome. Furthermore, the role of this circuit in different behaviors and associative memory permits the linking of patterns of receptor localization to circuit function and resulting behavior. Because of these features, the authors can provide evidence that two antagonizing dopamine receptors can act as autoreceptors within the axonal compartment of MB innervating DANs. The differential regulation of the balance of the two receptors under starvation in two distinct DAN innervations provides evidence of the role that regulation of this balance can play in circuit function and behavioral output.

Weaknesses:

The approach of using endogenously tagged alleles to study localization is a strength of this study, but the authors do not provide sufficient evidence that the insertion of 7 copies of split GFP to the C terminus of the dopamine receptors does not interfere with the endogenous localization pattern or function. Both sets of tagged alleles (1X Venus and 7X split GFP tagged) were previously reported (Kondo et al., 2020), but only the 1X Venus tagged alleles were further functionally validated in assays of olfactory appetitive memory. Despite the smaller size of the 7X split-GFP array tag knocked into the same location as the 1X venus tag, the reconstitution of 7 copies of GFP at the C terminus of the dopamine receptor, might substantially increase the molecular bulk at this site, potentially impeding the function of the receptor more significantly than the smaller, single Venus tag. The data presented by Kondo et al. 2020, is insufficient to conclude that the two alleles are equivalent.

In the revision, we validated the function of these engineered receptors by a new set of olfactory learning experiments. Both these receptors in KCs were shown to be required for aversive memory (Kim et al., 2007, Scholz-Kornehl et al., 2016). As in the anatomical experiments, we induced GFP110 expression in KC of the flies homozygous for 7xGFP11-tagged receptors using MB-Switch and 3 days of RU486 feeding o. We confirmed STM performance of these flies were not significantly different from the control (Figure 2 – figure supplement 1). Thus, these fusion receptors are functional.

The authors' conclusion that the receptors localize to presynaptic sites is weak. The analysis of the colocalization of the active zone marker Brp whole-brain staining with dopamine receptors labeled in specific neurons is insufficient to conclude that the receptors are localized at presynaptic sites. Given the highly crowded neuropil environment, the data cannot differentiate between the receptor localization postsynaptic to a dopamine release site or at a presynaptic site within the same neuron. The known distribution of presynaptic sites within the neurons analyzed in the study provides evidence that the receptors are enriched in axonal compartments, but co-labeling of presynaptic sites and receptors in the same neuron or super-resolution methods are needed to provide evidence of receptor localization at active zones. The data presented in Figures 5K-5L provides compelling evidence that the receptors localize to neuronal varicosities in DANs where the receptors could play a role as autoreceptors.

Given the highly crowded environment of the mushroom body neuropil, the analysis of dopamine receptor localization in Kenyon cells is not conclusive. The data is sufficient to conclude that the receptors are preferentially localizing to the axonal compartment of Kenyon cells, but co-localization with brain-wide Brp active zone immunostaining is not sufficient to determine if the receptor localizes juxtaposed to dopaminergic release sites, in proximity of release sites in Kenyon cells, or both.

To better resolve the microcircuits of KCs, we triple-labeled the plasma membrane and DAR::rGFP in KCs, and Brp, and examined their localizations with high-resolution imaging with Airyscan. This strategy revealed the receptor clusters associated with Brp accumulation within KCs (Figure 4). To further verify the association of DARs and active zones within KCs, we co-expressed Brpshort::mStraw and GFP1-10 and confirmed their colocalization (Figure 5A), suggesting presynaptic localization of DARs in KCs. With these additional characterizations, we now discuss the significance of receptors at the presynaptic sites of KCs.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

This is an important and interesting study that uses the split-GFP approach. Localization of receptors and correlating them to function is important in understanding the circuit basis of behavior.

For Figure 1, the authors show PAM, PPL1 neurons, and the ellipsoid body as a validation of their tools (Dop1R1-T2A-GAL4 and Dop2R-T2A-GAL4) and the idea that these receptors are colocalized. However, it appears that the technique was applied to the whole brain so it would be great to see the whole brain to understand how much labelling is specific and how stochastic. Methods could include how dissection conditions were controlled and how sensitive are receptor expression to the time of day of dissection, staining, etc.

The expression patterns of the receptor T2A-GAL4 lines (Figure 1A and 1B) are consistent in the multiple whole brains (Kondo et al., 2020, Author response image 1).

The significance of the expression of these two receptors in an active zone is not clearly discussed and presynaptic localization is not elaborated on. Would something like expansion microscopy be useful in resolving this? It would be important to discuss that as giant neurons in culture don't replicate many aspects of the MB system.

In the revised manuscript, we elaborated discussion regarding the function of the two antagonizing receptors at the AZ (Lines #226-275).

Does MB-GeneSwitch > GFP1-1 reliably express in gamma lobes? Most of the figures show alpha/beta lobes.

Yes. MB-GeneSwitch is also expressed in γ KCs, but weakly. 12 hours of RU486 feeding, which we did in the previous experiments, was insufficient to induce GFP reconstitution in the γ KCs. By extending the time of transgene induction, we visualized expression of Dop1R1 and Dop2R more clearly in γ KCs. Their localization is similar to that in the α/β KCs (Figure 4C, Figure 5 - figure supplement 1).

Figure 6, y-axis says protein level. At first, I thought it was related to starvation so maybe authors can be more specific as the protein level doesn't indicate any aspect of starvation.

We appreciate this comment, and the labels on the y-axis were now changed to “rGFP levels” (Figure 8C and 8F, Figure 8 - figure supplement 1B, 1D and 1F).

Reviewer #2 (Recommendations For The Authors):

Title:

The title of the manuscript focuses on the tagging of the receptors and their synaptic enrichment.

Given that the alleles used in the study were generated in a previously published study (Kondo et al, 2020), which describes the receptor tagging and that the data currently provided is insufficient to conclude that the receptors are localizing to synapses, the title should be changed to reflect the focus on localizing antagonistic cognate neurotransmitter receptors in the same neuron and their putative role as autoreceptors in DANs.

Following this advice, we removed the methodology from the title and revised it to “Synaptic enrichment and dynamic regulation of the two opposing dopamine receptors within the same neurons”.

Minor issues with text and figures:

Figure 1

A conclusion from Figure 1 is that the two receptors are co-expressed in Kenyon cells. Please provide panels equivalent to the ones shown in D-G, with Kenyon cells cell bodies, or mark these cells in the existing panels, if present. Line 111 refers to panel 1D as the Kenyon cells panel, which is currently a PAM panel.

We added images for coexpression of these receptors in the cell bodies of KCs (Figure 1 - figure supplement 1) and revised the text accordingly (Lines #89-90).

Given that most of the study centers on visualizing receptor localization, it would benefit the reader to include labels in Figure 1 that help understand that these panels reflect expression patterns rather than receptor localization. For instance, rCD2::GFP could be indicated in the Dop1R1-LexA panels.

As suggested, labels were added to indicate the UAS and lexAop markers (Figure 1D, 1E, 1G-1I and Figure 1 – figure supplement 1).

Given that panels D-E focus on the cell bodies of the neurons, it could be beneficial for the reader to present the ellipsoid body neurons using a similar view that only shows the cell bodies. Similarly, one could just show the glial cell bodies .

We now show the cell bodies of ring neurons (Figure 1G) and ensheathing glia (Figure 1I).

For panel 1E, please indicate the subset of PPL1 neurons that both expressed Dop1R1 and Dop2R, as indicated in the text, as it is currently unclear from the image.

Dop1R1-T2A-LexA was barely detected in all PPL1 (Figure 1E). We corrected the confusing text (Lines #95-96).

Figure 2

The cartoon of the cell-type-specific labeling should show that the tag is 7XFP-11 and the UAScomponent FP-10, as the current cartoon leads the reader to conclude that the receptors are tagged with a single copy of split GFP. The detail that the receptors are tagged with 7 copies of split GFP is only provided through the genotype of the allele in the resource table. This design aspect should be made clear in the figure and the text when describing the allele and approach used to tag receptors in specific neuron types.

We now added the construct design in the scheme (Figure 2A) and revised the corresponding text (Line #101-103).

Panel A. The arrow representing the endogenous promoter in the yellow gene representation should be placed at the beginning of the coding sequence. Currently, the different colors of what I assume are coding (yellow) and non-coding (white) transcript regions are not described in the legend. I would omit these or represent them in the same color as thinner boxes if the authors want to emphasize that the tag is inserted at the C terminus within the endogenous locus.

The color scheme was revised to be more consistent and intuitive (Figure 2A).

Figure 3

Labels of the calyx and MB lobes would benefit readers not as familiar with the system used in the study. In addition, it would be beneficial to the reader to indicate in panel A the location of the compartments analyzed in panel H (e.g., peduncle, α3).

Figure 3A was amended to clearly indicate the analyzed MB compartments.

Adding frontal and sagittal to panels B-E, as in Figure 2, would help the reader interpret the data.

In Figure 3B, “Frontal” and “Sagittal” were indicated.

Panel F-G. A scale bar should be provided for the data shown in the insets. Could the author comment on the localization of Dop1R1 in KCs? The data in the current panel suggests that only a subset of KCs express high levels of receptors in their axons, as a portion of the membrane is devoid of receptor signals. This would be in line with differential dopamine receptor expression in subsets of Kenyon cells, as shown in Kondo et al., 2020, which is currently not commented on in the paper.

We confirmed that the majority of the KCs express both Dop1R1 and Dop2R genes (Figure 1 - figure supplement 1). LIs should be compared within the same cells rather than the differences of protein levels between cell types as they also reflect the GAL4 expression levels.

Panel H. Some P values are shown as n.s. (p> 0.05). Other non-significant p values in this panel and in other figures throughout the paper are instead reported (e.g. peduncle P=0.164). For consistency, please report the values as n.s. as indicated in the methods for all non-significant tests in this panel and throughout the manuscript.

We now present the new dataset, and the graph represents the appropriate statistical results (Figure 3D; see the methods section for details).

The methods of labeling the receptors through the expression of the GeneSwitch-controlled GFP1-10 in Kenyon cells induced by RU486 are not provided in the methods. Please provide a description of this as referenced in the figure legend and the genotypes used in the analysis shown in the panels.

The method of RU486 feeding has been added. We apologize for the missing method.

Figure 4

Please provide scale bars for the inset in panels A-B.

Scale bars were added to all confocal images.

The current analysis cannot distinguish between postsynaptic and presynaptic dopamine receptors in KCs, and the figure title should reflect this.

We now present the new data dopamine receptors in KCs and clearly distinguish Brp clusters of the KCs and other cell types (Figure 4, Figure 5).

The reader could benefit from additional details of using the giant neuron model, as it is not commonly used, and it is not clear how to relate this to interpret the localization of dopaminergic receptors within Kenyon cells. The use of the venus-tagged receptor variant should be introduced in the text, as using a different allele currently lacks context. Figures 4F-4J show that the receptor is localizing throughout the neuron. Quantifying the fraction of receptor signal colocalizing with Brp could aid in interpreting the data. However, it would still not be clear how to interpret this data in the context of understanding the localization of the receptors in neurons within fly brain circuits. In the absence of additional data, the data provided in Figure 4 is inconclusive and could be omitted, keeping the focus of the study on the analysis of the two receptors in DANs. Co-expressing a presynaptic marker in Kenyon cells (e.g., by expressing Brp::SNAP) in conjunction with rGFP labeled receptor would provide additional evidence of the relationship of release sites in Kenyon cells and tagged dopamine receptors in these same cells and could add evidence in support to the current conclusion.

Following the advice, we added a short summary to recapitulate that the giant neurons exhibit many characteristics of mature neurons (Lines #152-156): "Importantly, these giant neurons exhibit characteristics of mature neurons, including firing patterns (Wu et al., 1990; Yao & Wu, 2001; Zhao & Wu, 1997) and acetylcholine release (Yao et al., 2000), both of which are regulated by cAMP and CaMKII signaling (Yao et al., 2000; Yao & Wu, 2001; Zhao & Wu, 1997)." Therefore, the giant neuron serves as an excellent model to study the presynaptic localization in large cells in isolation.

To clarify polarized localization of Brp clusters and dopamine receptors but not "localizing throughout the neuron", we now show less magnified data (Figure 5C). It clearly demonstrates punctate Brp accumulations localized to the axon terminals of the giant neurons (former Figure 4D and 4E). This is the same membrane segment where Dop1R1 and Dop2R are localized (Figure 5C). Therefore, the association of Brp clusters and the dopamine receptors in the isolated giant neurons suggests that the subcellular localization in the brain neurons is independent of the circuit context.

As the giant neurons do not form intermingled circuits, venus-tagged receptors are sufficient for this experiment and simpler in genetics.

Following the suggestion to clarify the AZ association of the receptors in KCs, we coexpressed Brpshort-mStraw and GFP1-10 in KCs and confirmed their colocalization (Figure 5A).

Figure 6

The data and analysis show that starvation induces changes in the α3 compartment in PPL1 neurons only, while the data provided shows no significant change for PPL1 neurons innervating other MB compartments. This should be clearly stated in lines 174-175, as it is implied that there is a difference in the analysis for compartments other than α3. Panel L of Figure 6 - supplement 1 shows no significant change for all three compartments analyzed and should be indicated as n.s. in all instances, as stated in the methods.

We revised the text to clarify that the starvation-induced differences of Dop2R expression were not significant (Lines #217-219). The reason to highlight the α3 compartment is that both Dop1R1 and Dop2R are coexpressed in this PPL1 neuron (Figure 8D).

Additional minor comments:

There are a few typos and errors throughout the manuscript. The text should be carefully proofread to correct these. Here are the ones that came to my attention:

Please reference all figure panels in the text. For instance, Figure 3A is not mentioned and should be revised in line 112 as Figure 3A-E.

Lines 103-104. The sentence "LI was visualized as the color of the membrane signals" is unclear and should be revised.

Figure 4 legend - dendritic claws should likely be B and C and not B and E.

Lines 147 - Incorrect figure panels, should be 5C-L or 5D-E.

Line 241 - DNAs should be DANs.

Methods - please define what the abbreviation CS stands for.

We really appreciate for careful reading of this reviewer. All these were corrected.

Version 2

Abstract

Abstract

Dopamine can play opposing physiological roles depending on the receptor subtype. In the fruit fly Drosophila melanogaster, Dop1R1 and Dop2R encode the D1- and D2-like receptors, respectively, and are reported to oppositely regulate intracellular cAMP levels. Here, we profiled the expression and subcellular localization of endogenous Dop1R1 and Dop2R in specific cell types in the mushroom body circuit. For cell-type-specific visualization of endogenous proteins, we employed reconstitution of split-GFP tagged to the receptor proteins. We detected dopamine receptors at both presynaptic and postsynaptic sites in multiple cell types. Quantitative analysis revealed enrichment of both receptors at the presynaptic sites, with Dop2R showing a greater degree of localization than Dop1R1. The presynaptic localization of Dop1R1 and Dop2R in dopamine neurons suggests dual feedback regulation as autoreceptors. Furthermore, we discovered a starvation-dependent, bidirectional modulation of the presynaptic receptor expression in the PAM and PPL1 clusters, two distinct subsets of dopamine neurons, suggesting regulation of appetitive behaviors. Our results highlight the significance of the co-expression of the two opposing dopamine receptors in the spatial and conditional regulation of dopamine responses in neurons.

Introduction

Neurotransmitters typically have multiple cognate receptors, and they may recruit different second messenger systems. Therefore, the expression and localization of receptor subtypes are critical for determining cellular responses to neurotransmitter inputs. The dopaminergic system offers an ideal in-vivo study case to this end, as it regulates a wide array of physiological functions through combinations of different receptor subtypes. In mammals, D1-like receptors are coupled with Gαs, thereby activating adenylate cyclase upon ligand binding, whereas Gαi-coupled D2-like receptors inhibit cyclase activity (Beaulieu & Gainetdinov, 2011). Four dopamine receptors have been identified in Drosophila: Dop1R1, Dop1R2, Dop2R, and DopEcR (K. A. Han et al., 1996; Hearn et al., 2002; Srivastava et al., 2005; Sugamori et al., 1995). Dop1R1 and Dop2R, also known as DopR1, dDA1 and dumb, and as DD2R, respectively, correspond to the D1- and D2-like receptors, respectively (Hearn et al., 2002; Sugamori et al., 1995). Dop1R2 and DopEcR are invertebrate-specific and have been reported to recruit different second messenger systems (K. A. Han et al., 1996; Srivastava et al., 2005). Intriguingly, recent data of single-cell RNA-seq and transgenic expression profiling revealed that the expression of these dopamine receptors is highly overlapping in the fly brain (Croset et al., 2018; Davie et al., 2018; Kondo et al., 2020), unlike the spatially segregated expression of the D1- and D2-like receptors in vertebrate brains (Gerfen & Surmeier, 2011). Considering the opposing physiological roles of the Dop1R1 and Dop2R, their protein localization, especially in those cells where they are co-expressed, should be critical to determine the responses to the dopamine inputs.

Drosophila mushroom bodies (MB) have long served as a unique dopaminergic circuit model to study adaptive behaviors, such as associative learning. MB-projecting neurons and their connections have been systematically described at both mesoscopic and ultrastructural resolutions (Aso, Hattori, et al., 2014; Li et al., 2020; Takemura et al., 2017; Tanaka et al., 2008). Kenyon cells (KCs), the major MB intrinsic neurons, encode a variety of sensory information (Honegger et al., 2011; Turner et al., 2008; Vogt et al., 2014). Intriguingly, each KC receives synaptic inputs from different dopaminergic projections in multiple spatially segmented compartments along its axon in the MB lobe (Aso, Hattori, et al., 2014; Tanaka et al., 2008). MB-projecting dopamine neurons (DANs) originate from the three cell clusters (Mao & Davis, 2009; Nässel & Elekes, 1992). DANs in the protocerebral posterior lateral 1 (PPL1) cluster project to the vertical lobes and the peduncle of the MB, and they control different aspects of associative memory (Aso et al., 2010, 2012; Aso & Rubin, 2016; Claridge-Chang et al., 2009; Krashes et al., 2009; Mao & Davis, 2009; Masek et al., 2015; Riemensperger et al., 2005; Takemura et al., 2017; Vogt et al., 2014). The protocerebral anterior medial (PAM) cluster, the largest DAN cluster, mostly projects to the medial lobes of the MB, and many PAM neurons are involved in reward processing (Burke et al., 2012; Felsenberg et al., 2018; Huetteroth et al., 2015; Ichinose et al., 2021; Lin et al., 2014; Liu et al., 2012; Yamagata et al., 2015, 2016). DANs in PPL2ab project to the MB calyx and control the conditioned odor response (Boto et al., 2019). In addition to the variety of the dopamine sources, KCs express all four dopamine receptor subtypes (Deng et al., 2019; Kondo et al., 2020). Besides KCs, these lobe-projecting DANs have synaptic outputs to multiple types of neurons, including MBONs, APL and DPM (Li et al., 2020; Takemura et al., 2017; Zhou et al., 2019). Given the multitude of modulatory effects of dopamine in the MB (Berry et al., 2018; Cohn et al., 2015; Handler et al., 2019), receptor localization in each cell type provides important information for interpreting such functional diversity.

The projections and synapses of the neurons in the MB circuit are tightly intertwined (Li et al., 2020; Takemura et al., 2017). Therefore, conventional immunohistochemical approaches using light microscopy do not allow identification of cells from which immunoreactive signals originate. Precise determination of their subcellular localization requires conditional visualization of the proteins of interest only in the target MB neurons (Bonanno et al., 2024; Fendl et al., 2020; Sanfilippo et al., 2024). Employing the CRISPR/Cas9-mediated split-GFP tagging strategy (D. Kamiyama et al., 2016; R. Kamiyama et al., 2021; Kondo et al., 2020), we profiled the spatial distribution of endogenous Dop1R1 and Dop2R proteins in KCs, the PAM, and the PPL1 DANs.

Results

Co-expression of Dop1R1 and Dop2R genes in the adult Drosophila brain

To compare the expression of different dopamine receptor genes in detail, we used T2A-GAL4 knock-ins of the endogenous Dop1R1 and Dop2R genes (Kondo et al., 2020) with fluorescent reporters. Both lines labelled many neuropils including the MB (Figure 1A and 1B), and the overlapping expression of these receptors is consistent with our previous quantification of GAL4-positive cells for both genes (58,049 and 68,528 for Dop1R1 and Dop2R, respectively, out of 118,331 brain cells; Kondo et al., 2020). Double labelling of Dop1R1-T2A-LexA and Dop2R-T2A-GAL4 expression revealed cells with overlapping and differential patterns (Figure 1C-1I; see also Kondo et al., 2020). We confirmed the co-expression of Dop1R1 and Dop2R in the KCs (Figure 1 – figure supplement 1) as reported previously (Kondo et al., 2020). The PAM cluster of DANs expressed both Dop1R1 and Dop2R (Figure 1C and 1D). On the other hand, most of the DANs in the PPL1 cluster strongly expressed Dop2R, but Dop1R1 only weakly (Figure 1C and 1E). We further found that Dop1R1, but not Dop2R, was highly expressed in the ring neurons projecting to the ellipsoid body (Figure 1G and 1H; Hanesch et al., 1989) and the neuropil ensheathing glia (Figure 1I; Awasaki et al., 2008). In conclusion, Dop1R1 and Dop2R genes are co-expressed in PAM neurons and KCs but have selective expressions in other cell types.

Co-expression of Dop1R1 and Dop2R genes in adult Drosophila brain.

(A and B) The expression of Dop1R1-T2A-GAL4 and Dop2R-T2A-GAL4 visualized by UAS-mCD8::GFP. Maximum-intensity projections of the whole brain.

(C) Schematic of the KCs and the MB-innervating dopamine neurons from the PAM and PPL1 clusters.

(D-E, G-I) Double labelling of Dop1R1-T2A-LexA and Dop2R-T2A-GAL4 expressions by lexAop-rCD2::GFP (green) and UAS-CD4::tdTomato (red), respectively. Dopamine neurons were immunostained with anti-TH antibody (blue). Single optical sections are shown. Cell bodies of the PAM cluster (D), the PPL1 cluster (E), ring neurons projecting to the ellipsoid body (G and H) and ensheathing glia (I) are shown.

(F) Schematic of the regions shown in (G-I).

Scale bars, 50 µm (A and B), 5 µm (D, E and I), 20 µm (G and H).

Quantification of subcellular enrichment of endogenous proteins in target cells

The overlapping expression of Dop1R1 and Dop2R genes prompted us to examine the subcellular localization of these receptor proteins. To elucidate the localization of these broadly expressed receptors (Figure 1), we took advantage of split-GFP tagging of endogenous proteins (D. Kamiyama et al., 2016; Kondo et al., 2020). By adding seven copies of GFP11 tags to the C-termini of the Dop1R1 and Dop2R proteins, their intracellular distribution can be visualized specifically in cells expressing GFP1-10 through split-GFP reconstitution (Figure 2A). To verify the functional integrity of reconstituted GFP (rGFP)-fused receptors, we examined aversive olfactory memory of homozygous flies carrying GFP11-tagged dopamine receptors and induced GFP reconstitution in KCs. Both Dop1R1 and Dop2R have been shown to be required for aversive memory in α/β and γ KCs (Kim et al., 2007; Scholz-Kornehl & Schwärzel, 2016). Memory scores of the rGFP-fused receptor flies were comparable to that of control flies without GFP11 insertion (Figure 2 – figure supplement 1).

Cell type-specific visualization of endogenous proteins with GFP11 tag.

(A) Principle of cell type specific fluorescent labelling of target proteins by GFP11 tag. Seven copies of GFP11 are fused to the C-terminal of endogenous receptors. GFP1-10 and membrane marker CD4::tdTomato are expressed in the target cells by GAL4/UAS system. In the target cells, reconstitution of GFP occurs on the endogenous proteins tagged with GFP11.

(B) As an example, DopEcR::GFP11 is visualized in KCs using MB-Switch, a ligand-inducible GAL4 driver. To activate Gene-Switch, flies were fed with food containing 200 µM RU486 for 12 hours before dissection. A merged image of reconstituted GFP (green) and cellular membrane visualized by CD4::tdTomato (magenta). Maximum intensity projection of the whole left MB.

(C) The workflow for visualizing subcellular protein enrichment by localization index (LI). A single sagittal section of the MB calyx and peduncle is shown. The ratio of reconstituted GFP to membrane signal is calculated and normalized by the mean of all voxels to provide LI. In the middle image, LI is color-coded so that red represents local receptor enrichment. In the right image, the intensity of LI color is adjusted based on the membrane signal.

To quantify the subcellular enrichment of the receptors, we devised Localization Index (LI). Briefly, LI is the normalized ratio of the rGFP signal to the reference membrane marker signal (CD4::tdTomato). If the target and reference proteins had the identical distribution, LI would be 1 everywhere. More or less enrichment of rGFP would result in the LI larger or smaller than 1, respectively (Figure 2C; see also Methods for detail). For visualization, the reference membrane signal was color-coded with LI (Figure 2C). As proof of principle, mapping the LI of DopEcR::rGFP signals in KCs highlighted enrichment in the proximal peduncle (Figure 2B and 2C), which is consistent with the previous report (Kondo et al., 2020). This representation thus visualizes the subcellular enrichment of the targeted receptors in the plasma membrane of GAL4-expressing cells.

Colocalization of Dop1R1 and Dop2R proteins

First, we compared the localization of Dop1R1 and Dop2R proteins in KCs, where both receptor genes were highly expressed (Figure 1 – figure supplement 1). These receptors were distributed throughout KC membranes with predominant enrichment in the lobes (Figure 3A-3C), but sparsely in the calyx (Figure 3B and 3C). LI quantification revealed that enrichment in the lobes was more pronounced in Dop2R compared to Dop1R1, whereas localization to the calyx was sparser in Dop2R than Dop1R1 (Figure 3D). We confirmed that the differential localization of these receptors was consistent across multiple experimental batches conducted on different days (Figure 3 – figure supplement 1).

Subcellular localization of Dop1R1 and Dop2R in the Kenyon cells.

Subcellular localization of Dop1R1 and Dop2R in the KCs is visualized by GFP11-tag. MB-Switch was used to express GFP1-10 and CD4::tdTomato in the KCs. To activate Gene-Switch, flies were fed with food containing 200 µM RU486 for 72 hours before dissection.

(A) Schematic showing the projection pattern of a α/β KC.

(B) Enrichment of Dop1R1 and Dop2R in the MB lobe. Maximum-intensity projections of the lobe (left) and the calyx (middle) are shown in frontal view. The whole left MB are shown in sagittal view (right). Reconstituted GFP signals for both Dop1R1:: and Dop2R::GFP11 distributed throughout the MB lobe and the calyx.

(C) Visualization by LI showed more pronounced enrichment of Dop2R than Dop1R1 in the lobe.

(D) Mean LI of Dop1R1 and Dop2R in the calyx, the peduncle, the α3 and β2 compartment in the lobe. Student’s t test was performed to compare LI of Dop1R1 and Dop2R in each region (N = 10). Error bars; SEM. p> 0.05, ** p< 0.01, *** p< 0.001, **** p< 0.0001, ns: not significant p>0.05.

Scalebars, 20 µm (B and C).

KCs have major presynaptic sites in the MB lobes, where Dop1R1 and Dop2R are enriched (Figure 3A-3D). Therefore, we examined receptor localization with the reference of Brp immunostaining, which labels the active zones (AZ) (Wagh et al., 2006). To distinguish the AZ in KCs from those in the other neurons, we co-labeled the plasma membrane of KCs and conducted high-magnification imaging using Airyscan. At this resolution, Brp puncta in KCs and those from non-KCs could be distinguished based on their overlap with the KC membrane marker (Figure 4B and 4C). Interestingly, KC-specific visualization of Dop1R1 and Dop2R proteins revealed signals around the Brp puncta of the same cells, suggesting presynaptic localization (Figure 4B and 4C). Additionally, we found the receptor condensates adjacent to the Brp clusters of non-KCs, suggesting their localization at the postsynaptic sites (Figure 4B and 4C). The synaptic receptor localization is consistent in different subsets of KCs (α/β and γ; Figure 4A-4C). In conclusion, both of these antagonizing dopamine receptors are enriched in the presynaptic and postsynaptic sites of KCs (Figure 4D).

High-resolution imaging revealed the two opposing dopamine receptors existing on the presynaptic and postsynaptic sites of Kenyon cells.

(A) Schematic showing the projection pattern of α/β and γ KCs.

(B and C) Airyscan images of Dop1R1::rGFP and Dop2R::rGFP in KCs (green) co-labeled with the AZ stained with anti-Brp (magenta). MB-Switch was used with 72 hours of RU486 feeding to express GFP1-10 and CD4::tdTomato in the KCs. Brp puncta that overlap with CD4::tdTomato signals (cyan) are identified to be presynaptic sites of in KCs (white arrowheads), and those do not overlap are determined to be presynaptic sites of non-KCs (outlined arrows). The synaptic localization of these receptors is similar in the α 3 (B) and γ5 (C) compartments. In the right panels, white squared regions in the left panels are magnified. Scalebars, 5 µm (left), 1 µm (right).

(D) Illustration of localization of Dop1R1 and Dop2R to presynaptic and postsynaptic sites in the axon terminal of KCs.

To further clarify the presynaptic localization in KCs, we labeled the AZ of KCs by expressing Brpshort::mStraw (Fouquet et al., 2009) and confirmed that both Dop1R1::rGFP and Dop2R::rGFP were associated with the Brp puncta in the lobes (Figure 5A, Figure 5 – figure supplement 1). Interestingly, we found that not all Brp puncta of KCs were associated with the dopamine receptors (Figure 5A), suggesting that dopaminergic presynaptic modulation is heterogeneous across release sites. This heterogeneity can well explain differential learning-induced plasticity across boutons within single KCs (Bilz et al., 2020; Davidson et al., 2023).

Presynaptic localization of Dop1R1 and Dop2R in Kenyon cells and giant neurons.

(A) Double labelling of dopamine receptors (green) and the AZ of the KCs (magenta). MB-Switch was used with 72 hours of RU486 feeding to express GFP1-10 and Brpshort::mStraw in the KCs. Single focal slices at the α3 compartment are shown. White squares in the left panels are magnified in the right panels. The Brp puncta in KCs were either abutted by the dopamine receptor signals (white arrowheads) or had barely detectable signals nearby (outlined arrowheads). Scale bars, 5 μm (left), 1 μm (right).

(B) Punctate Brp expression in a giant neuron culture differentiated from cytokinesis-arrested neuroblasts of OK371-GAL4/UAS-mCD8::GFP embryos. Aggregated Brp condensates (magenta) were observed in the neurite terminals of the cells marked with mCD8::GFP (cyan) in the right panel. Scale bars, 20 µm (left), 10 μm (right).

(C) Double labelling of dopamine receptors (green) and the AZs (magenta). Dop1R1::Venus or Dop2R::Venus was crossed with Brp::SNAP. In the left panels, giant neurons extending their neurites from the cell body on the left to the right. In the right panels, white squared regions in the left panels are magnified. Scale bars, 10 μm (left), 2 μm (right).

To better resolve the presynaptic localization of Dop1R1 and Dop2R, we turned to “giant” Drosophila neurons differentiated from cytokinesis-arrested neuroblasts in culture (Wu et al., 1990). The expanded size of the giant neurons is advantageous for investigating the microanatomy of neurons in isolation (Saito & Wu, 1991; Wu et al., 1990; Yao et al., 2000). Importantly, these giant neurons exhibit characteristics of mature neurons, including firing patterns (Wu et al., 1990; Yao & Wu, 2001; Zhao & Wu, 1997) and acetylcholine release (Yao et al., 2000), both of which are regulated by cAMP and CaMKII signaling (Yao et al., 2000; Yao & Wu, 2001; Zhao & Wu, 1997). In the giant neurons from the Brp::SNAP embryos (Kohl et al., 2014), Brp was localized to the terminals of neurites, but rarely in the proximal neurites (Figure 5B). Furthermore, we found punctate Brp clusters in the giant neuron terminals (Figure 5B), together recapitulating the essential characteristics of the AZ cytomatrix in adult neurons. To confirm if dopamine receptors localize to these presynaptic sites, we generated the giant neurons from embryos carrying the Venus insertion to Dop1R1 and Dop2R (Kondo et al., 2020) together with Brp::SNAP. Both Dop1R1 and Dop2R were expressed in the giant neurons and enriched in the same distal axonal segments (Figure 5C). A closer investigation revealed that these receptors are associated with the Brp clusters (Figure 5C). These observations in the giant neurons are strikingly similar to those in the KCs (Figure 3, 4 and 5A), corroborating that the presynaptic localization of these receptors is independent of the circuit context.

Distinct subcellular enrichment of dopamine receptors in MBONs and DANs

Presynaptic and postsynaptic localization of Dop1R1 and Dop2R in KCs prompted us to investigate MBONs, that have profound postsynaptic contacts with DANs in the MB compartments (Li et al., 2020; Takemura et al., 2017). MBON-γ1pedc>αβ (also known as MB-MVP2, MBON-11; Aso, Hattori, et al., 2014; Tanaka et al., 2008) has most of its postsynaptic sites on the γ1 compartment and the peduncle of the α/β neurons and send axonal projections to the α and β lobes (Figure 6; Dorkenwald et al., 2024; Schlegel et al., 2024). We analyzed the subcellular localization of Dop1R1::rGFP and Dop2R::rGFP in MBON-γ1pedc>αβ by driving GFP1-10 using R83A12-GAL4 (Perisse et al., 2016). Strikingly, both Dop1R1 and Dop2R are enriched in the dendritic projection in the γ1 compartment (Figure 6B and 6C) in sharp contrast to the sparse distribution of these receptors in the dendrites of KCs (Figure 3B-3D). This localization is consistent with previous reports of functional DAN>MBON synapses (Takemura et al., 2017) and dopaminergic plasticity on the dendrites (Boto et al., 2019; Pribbenow et al., 2022). Additionally, we detected these receptors in the presynaptic boutons (Figure 6C), which are consistent with the results in KCs and in the giant neurons (Figure 4B, 4C, 5A and 5C).

Subcellular localization of Dop1R1 and Dop2R in MBON-γ1pedc>αβ.

(A) The projection pattern of MBON-γ1pedc>αβ from the tracing data in FlyWire (Dorkenwald et al., 2024; Schlegel et al., 2024).

(B and C) R83A12-GAL4 was used to express UAS-GFP1-10 and UAS-CD4::tdTomato in the PAM neurons.

(B) Reconstituted GFP signals (left) of Dop1R1::GFP11 and Dop2R::GFP11 in MBON-γ1pedc>αβ. Maximum-intensity projections of the left MB lobe. Visualization of LI (right) revealing that both Dop1R1 and Dop2R are enriched in the dendritic projection of MBON-γ1pedc>αβ in the γ1 compartment as well as in the presynaptic boutons.

(C) Airyscan images of the presynaptic boutons aroundα3 (left) and dendritic projections in the γ1 compartment (right). White squares in the right panels are magnified in the insertion to show the swelling membrane structures with punctate localization of dopamine receptors.

Scale bars, 20 µm (B), 5 µm (C), 1 µm (C, insertion).

D2-like receptors in mammals are expressed in DANs and act as autoreceptors, which mediates feedback signals by receiving neurotransmitters released from the neuron on which the receptors reside (Ford, 2014). In Drosophila, multiple dopamine receptor genes are expressed in DANs (Figure 1D and 1E; Aso et al., 2019; Deng et al., 2019; Kondo et al., 2020), but it is unclear if these receptors function as autoreceptors. We therefore examined the subcellular localization of Dop1R1 and Dop2R in the PAM cluster DANs with a particular focus on their presynaptic terminals. The PAM neurons are polarized; presynaptic proteins are abundantly enriched in the MB lobe and barely detected in dendrites (Figure 7A and 7B). We visualized Dop1R1 and Dop2R proteins in the PAM cluster DANs using R58E02-GAL4, and both were localized to the terminals in the MB lobes (Figure 7C). Representation of LI for Dop1R1 and Dop2R in the PAM neurons again showed stronger presynaptic enrichment of Dop2R than Dop1R1 (Figure 7C). Dop2R was strongly enriched at β’1 compartment showing significantly higher LI than Dop1R1 (Figure 7D and 7E). Higher magnification revealed the accumulation of both Dop1R1 and Dop2R in the boutons (Figure 7F). The presynaptic localization of Dop1R1 and Dop2R in DANs suggests that both receptors mediate feedback regulation.

Subcellular localization of Dop1R1 and Dop2R in dopamine neurons.

(A) Maximum-intensity projection image showing the distribution of presynaptic sites in the PAM neurons. Left panel: R58E02-GAL4 was used to express mCD8::GFP (magenta) and nSyb::CLIP (magenta). Right panel: Visualization by LI showing enrichment of nSyb signals in the lobe projection of the PAM neurons.

(B) Illustrated projection pattern of the PAM neurons. Red puncta on the dendrites indicate the sparse distribution of presynaptic sites in dendrites.

(C-F) Subcellular localization of GFP11-tagged Dop1R1 and Dop2R in the PAM neurons. R58E02-GAL4 (C and F) or R15A04-GAL4 (D and E) was used to express UAS-GFP1-10 and UAS-CD4::tdTomato in the PAM neurons.

(C) Reconstituted GFP signals of Dop1R1::GFP11 and Dop2R::GFP11 in PAM neurons (left). LI visualization revealed the stronger presynaptic enrichment of Dop2R than that of Dop1R1 (right). Maximum-intensity projections of the left hemisphere including the whole MB lobe and dendritic projections of the PAM neurons around the MB.

(D and E) LI in PAM-β’1 neuron. (D) The presynaptic terminals of PAM-β’1 neurons are shown (dashed line). (E) Mean LI for Dop1R1 and Dop2R in the β’1 (Mann-Whitney U test, N = 9). Error bars; SEM.

(F) A single optical slice of the γ5 compartment in the MB lobe obtained using Airyscan. Merged image of reconstituted GFP (green) and CD4::tdTomato (magenta). Insertions are the magnified images of the presynaptic boutons of PAM-γ5 (white squares).

Scale bars, 20 µm (A, C, D and F), 5 µm (F, insertion).

State-dependent and bidirectional modulation of dopamine receptor expression in the PAM and PPL1 DANs

The activity of MB-projecting DANs is reported to be dynamic and sensitive to feeding states (Ichinose et al., 2017; Liu et al., 2012; Plaçais & Preat, 2013; Senapati et al., 2019; Siju et al., 2020; Tsao et al., 2018; Yamagata et al., 2016). We therefore examined if starvation alters the protein expression of Dop1R1 and Dop2R in these MB-projecting DANs (Figure 8A). We found a significant elevation of Dop1R1 in the terminals of PAM-γ5 upon starvation for 10 hours or longer (Figure 8B and 8C). In contrast, starvation did not increase Dop2R in PAM-γ5 but rather tended to decrease, if at all (Figure 8B and 8C). We found similar starvation-dependent changes in Dop1R1 and Dop2R levels in other PAM neurons (Figure 8 – figure supplement 1A-1D). These results together suggest that starvation enhances presynaptic dopamine signaling in the reward-related PAM neurons by shifting the balance of Dop1R1 and Dop2R.

Bidirectional modification of dopamine receptor expression in dopamine neurons.

(A) Schematic illustration of the MB projection of the PAM and PPL1 dopamine neurons.

(B) Dop1R and Dop2R in the presynaptic terminals of PAM-γ5 after 48 hours of starvation compared with fed state.

(C) Quantification of rGFP signal levels in the presynaptic terminals of PAM-γ5 after 0, 10, 24 and 48 hours of starvation (n = 6-13).

(D) Reconstituted GFP signals of Dop1R1::GFP11 and Dop2R::GFP11 in the PPL1 neurons. In the MB projections of the PPL1 neurons, Dop1R1 was detected in only the α3 compartment. Dop2R was found in all MB projections. Maximum-intensity projections of the MB lobe.

(E) Dop1R and Dop2R in the presynaptic terminals of PPL1-α3 after 24 hours of starvation compared with fed state.

(F) Quantification of rGFP signal levels in the presynaptic terminals of PPL1-α3 after 0, 10 and 24 hours of starvation (n = 7-10).

Scale bar, 10 µm (B and E), 20 µm (D).

Interaction effects between genotypes and starvation time on protein levels were tested by Two-way ANOVA (C and F). Bars and error bars represent mean and SEM, respectively (C and F). ** p< 0.01, *** p< 0.001, ns: not significant p>0.05.

DANs of the PAM and PPL1 clusters exert distinct, largely opposite behavioral functions (Claridge-Chang et al., 2009; Liu et al., 2012). Therefore, plasticity in the PPL1 neurons may differ from that in the PAM neurons. To test this hypothesis, we examined the starvation-dependent change in Dop1R1 and Dop2R protein expression in the PPL1 neurons. To this end, we visualized GFP11-tagged receptors by expressing GFP1-10 using TH-GAL4 (Friggi-Grelin et al., 2003). We detected Dop2R proteins in all MB projections of the PPL1 neurons, whereas Dop1R1 proteins were only detectable in the terminals of the PPL1-α3 neuron (Figure 8D). Strikingly, the starvation-induced changes in the PPL1 neurons were opposite to those in the PAM: Dop2R, but not Dop1R1, was significantly increased in the α3 compartment (Figure 8E and 8F). In the other compartments, the Dop2R::rGFP levels tended to be higher in starved flies although the increase was not statistically significant (Figure 8 – figure supplement 1E and 1F). These results are in line with the state-dependent changes in the physiology of these DANs (Siju et al., 2020; Tsao et al., 2018). Taken together, starvation induces bidirectional modulation of the dual autoreceptor system in the PPL1 and PAM DANs, and we propose that these changes shift the balance of dopamine output from these subsets (Figure 9A).

The dual dopaminergic feedback regulating starved-state dependent expression of appetitive behavior.

(A) A working model showing the role of the dual dopaminergic feedback regulation. In the starved state, increased Dop1R1 in PAM neurons and increased Dop2R in PPL1 neurons changes the balance between the synaptic outputs from these DANs to favor appetitive behavior.

(B) According to the model, loss of Dop2R in PPL1 upregulates output from PPL1 to attenuate appetitive behavior in starved flies.

(C) Knockdown of Dop2R in the PPL1 neurons by MB504B-GAL4 reduced 3-hour appetitive memory performance (t-test with Bonferroni correction, n=14-15). Bars and error bars represent mean and SEM, respectively. * p< 0.05.

The dopaminergic signals from the PPL1 neurons inhibit expression of appetitive olfactory memory in fed flies (Krashes et al., 2009; Pavlowsky et al., 2018). The increased Dop2R autoreceptors in starved flies (Figure 8E and 8F) may thus disinhibit appetitive memory expression by suppressing PPL1 outputs (Figure 9B). To test this hypothesis, we examined appetitive memory by transgenic knockdown of Dop2R specifically in the PPL1 cluster neurons using MB504B-GAL4 (Vogt et al., 2014). Indeed, appetitive memory was impaired (Figure 9C), suggesting the significance of the negative feedback regulation of the PPL1 via Dop2R in appetitive behavior expression. This result is consistent with our model, which suggests that enhanced negative feedback to PPL1 via Dop2R in the starved state contributes to the expression of appetitive behavior (Figure 9A and 9B).

Discussion

The present study demonstrated the expression, subcellular localization, and dynamics of endogenous dopamine receptors, specifically Dop1R1 and Dop2R, in the MB circuit of the fly brain. Compared to intronic insertions of reporters using transposons (Fendl et al., 2020), CRISPR/Cas9-based insertion is advantageous in terms of flexibility in fusion sites (Bonanno et al., 2024; R. Kamiyama et al., 2021; Sanfilippo et al., 2024; Williams et al., 2019). This enables the C-terminal tagging of dopamine receptors that do not perturb protein localization and function (Figure 2 – figure supplement 1).

The split-GFP reconstitution strategy enables sensitive detection of low-abundant endogenous proteins by the tandem multimerization of the GFP11 tags, as shown in other epitope tags such as SunTag and smFPs (Tanenbaum et al., 2014; Viswanathan et al., 2015). The native fluorescence of rGFP was sufficient to detect tagged proteins at the high signal-to-noise ratio, without the need for signal amplification using antibodies. The background fluorescence of the split-GFP fragments was practically negligible (Figure 8D; D. Kamiyama et al., 2016; Kondo et al., 2020). This approach can thus be applicable to monitor localization and dynamics of endogenous proteins of low abundance.

Our cell-type-specific receptor labeling offers further potential for ultrastructural examination using higher resolution imaging techniques. For example, synaptic condensates of Dop1R1 and Dop2R at the resolution in this study (Figures 4 and 5) did not allow us to distinguish the receptors on the plasma membrane and those that are internalized (Dumartin et al., 1998; Kotowski et al., 2011). Thus, application of super-resolution fluorescent imaging and expansion microscopy (Gao et al., 2019) would disentangle the precise membrane localization of the receptors.

Spatial regulation of dopamine signaling through the two opposing receptors

How do Dop1R1 and Dop2R function in presynaptic terminals? Released dopamine is not confined to the synaptic cleft, but diffused in the extracellular space (Liu et al., 2021; Rice & Cragg, 2008). Such volume transmission can increase the number of target synapses by 10 times (Li et al., 2020; Takemura et al., 2017). Thus, the broader distribution of Dop1R1 (Figure 3B-3D and 7C-7E) may enable it to respond to extrasynaptic dopamine. In response to residual dopamine, Dop2R at the AZ may inhibit activities of adenylate cyclases and voltage-gated calcium channels (Beaulieu & Gainetdinov, 2011; Hearn et al., 2002), thereby reducing the noise of the second messengers. Collectively, the spatial configuration of Dop1R1 and Dop2R may enhance the sensitivity and precision of dopaminergic modulation.

Furthermore, presynaptic receptor localization (Figure 4B-4D, 5A, 5C, 6C and 7F) suggests a spatially confined cAMP, forming nanodomains (Anton et al., 2022; Bock et al., 2020; Maiellaro et al., 2016; Zhang et al., 2020). cAMP signaling has been shown to regulate multiple events at presynaptic terminals, such as the molecular assembly at the AZ and synaptic vesicle dynamics (Ehmann et al., 2018; Kittel & Heckmann, 2016; Kuromi & Kidokoro, 2000, 2005; Maiellaro et al., 2016; Renger et al., 2000; Sachidanandan et al., 2023). Especially in KCs, cAMP-dependent plasticity underlies associative memory (Boto et al., 2014; Cohn et al., 2015; Louis et al., 2018; Noyes & Davis, 2023; Stahl et al., 2022). Presynaptic cAMP regulation through Dop1R1 and Dop2R therefore explains the requirement of these receptors in associative memory (Kim et al., 2007; Scholz-Kornehl & Schwärzel, 2016).

The dual autoreceptor system may shape dopamine release

Presynaptic localization of Dop1R1 and Dop2R in DANs (Figure 7C-7F) strongly suggests their functions as autoreceptors. In the Drosophila nervous system, Dop2R was shown to negatively regulate dopamine release (Shin & Venton, 2022; Vickrey & Venton, 2011). Tight presynaptic enrichment in PAM (Figure 7F) suggests that Dop2R receives high levels of dopamine and effectively prevents overactivation. The presence of Dop1R1 in DAN terminals (Figure 7F) was unexpected and introduces another layer of presynaptic regulations to dopamine release. Although it must be functionally verified, presynaptic Dop1R1 likely provides positive feedback to the dopamine release as an autoreceptor. This positive feedback would be particularly important for signal amplification when extracellular dopamine concentrations are low. Consistent this hypothesis, presynaptic Dop1R1 was undetectable in most PPL1 DANs (Figure 8D), which were reported to have high spontaneous activities (Feng et al., 2021; Plaçais & Preat, 2013). Taken together, this dual autoreceptor system likely fine-tune the amplitude and kinetics of dopamine release. Alternatively, these presynaptic receptors could potentially receive extrasynaptic dopamine released from other DANs. Therefore, the autoreceptor functions need to be experimentally clarified by manipulating the receptor expression in DANs.

Dopamine receptor expression is reported to be associated with prolonged exposure to psychoactive substances, such as caffeine and ethanol (Andretic et al., 2008; Kanno et al., 2021; Kondo et al., 2020; Petruccelli et al., 2018). Our study further showed starvation-dependent changes of Dop1R1 and Dop2R in DAN terminals (Figure 8). Strikingly, starvation responses of presynaptic Dop1R1 and Dop2R were differential depending on the DAN cell types (Figure 8). These results indicate that starvation bidirectionally changes the dual autoreceptor system, putting more weight on the PAM output over PPL1 to control the expression of appetitive behavior (Figure 9). This shifted balance explains the state-dependent changes in the presynaptic activity of these two clusters of DANs (Siju et al., 2020; Tsao et al., 2018) and bidirectional modulation of the MB output (Aso, Sitaraman, et al., 2014; Ichinose et al., 2021; Owald et al., 2015).

Materials and Methods

Key resource table

Flies

Flies were raised on standard cornmeal food at 25°C under a 12:12 h light-dark cycle (ZT0 at 8 AM) for all experiments. The GAL4-UAS system was used to express the transgenes of interest in specific neuron subtypes. Flies carrying GAL4 were crossed to another strain carrying UAS reporters, and F1 progenies were used in the experiments. To visualize GFP11-tagged dopamine receptors in the specific cell-types, female fly strains carrying UAS-CD4::tdTomato, UAS-GFP1-10 and GAL4 driver were crossed to male fly strains carrying Dop1R1::GFP11 or Dop2R::GFP11, and F1 progenies were used. To make the giant neuron culture, embryos from Brp::SNAP or the F1 progeny of OK371-GAL4 crossed with Brp::SNAP, UAS-mCD8::GFP, UAS-nSyb::CLIP flies were used. For the fly lines used in our manuscript, see Key resource table.

RU486 feeding

To activate Gene-Switch, flies are fed with food containing RU486 (mifepristone). Food containing 200 µM of RU486 was prepared as described previously (Mao et al., 2004). In brief, 200 mg of RU486 was dissolved in 10 ml of 99.5% ethanol to make a stock solution. 4.3 µl of the stock solution was added to 1 ml of molten fly food and mixed thoroughly. The molten food with RU486 was poured into vials or bottles and cooled to make solid food.

Brain dissection and Immunohistochemistry

Flies were sorted under CO2 anesthesia to select males with the specific genotypes and kept in a food vial for recovery at least 24 h prior to the dissection. Fly brains were dissected 3-7 days after eclosion in ice-cold Phosphate-buffered saline (PBS). After dissection, brains were kept in ice-cold PBS with 2% paraformaldehyde (PFA) for up to 30 min. For fixation, brains were incubated in 2 % PFA in PBS for 1 hour at room temperature. Fixed brains were washed in PBS with 0.3 % Triton X-100 (PBT) for 10 min 3 times. For rGFP imaging, fixed brains were mounted in SeeDB2 (Ke et al., 2016), and native fluorescence was imaged without signal amplification.

For chemical tagging reaction of nSyb::CLIP, brains were incubated in PBT containing CLIP-Surface 547 substrate (1µM) for 15 min at room temperature. Brains were washed in PBT for 10 min 3 times.

For immunohistochemistry, fixed brains were blocked in 3 % normal goat serum (NGS; Sigma-Aldrich, G9023) for 30 min at room temperature. Brains were incubated in primary and secondary antibodies diluted in PBT with 1 % NGS over two nights at 4 ℃, respectively. After each step, brains were washed three times in PBT for longer than 20 min at room temperature and mounted in 86% glycerol. The following primary antibodies were used: mouse anti-TH (1:100), mouse anti-Brp (1:40), rabbit anti-DsRed (1:2000). Secondary antibodies: Alexa Fluor 633 goat anti-mouse (1:400 for anti-TH, 1:200 for anti-Brp).

Fluorescent imaging

For image acquisition, whole-mount brains were scanned with the Olympus FV1200 confocal microscope with following objective lens; 20x oil (NA = 0.85, UPLSAPO20XO, Olympus; Figure 1A, 1B, 1G and 1H), 30x silicone (NA = 1.05, UPLSAPO30XS, Olympus; Figure 3B and 3C), 40x oil (NA = 1.3, UPLFLN40XO, Olympus; Figure 2B, 2C, 6B, 7A, 7C, 7D, 8B, 8E, Figure 8 – figure supplement 1-A,C,E), or 60x oil (NA = 1.42, PLAPON60XO, Olympus; Figure 1D, 1E, 1I, Figure 1 – figure supplement 1). Z-stack images were acquired.

For high resolution imaging in Figure 4B, 4C, 5A, 6C, 7F and Figure 5 – figure supplement 1, we used Airyscan on Zeiss LSM800 with 63x oil objective lens (NA = 1.40, Plan-Apochromat).

Image analysis

All image analysis were conducted on Fiji (Schindelin et al., 2012).

To visualize the subcellular localization of dopamine receptors in defined neurons, we devised the localization index (LI). Relative receptor density was calculated in each voxel by dividing the receptor (reconstituted GFP) signals by the corresponding membrane signals (CD4::tdTomato). To set ROI, any voxels devoid of membrane signal were censored. The local density was normalized by the mean values in the ROI. For the visual representation, the reference membrane signals were colored according to the normalized LI (Figure 2C). To quantify mean LI in each subcellular region (Figure 3H, 7E), ROI was manually set.

Embryonic giant neuron culture

Multinucleated giant neurons from neuroblasts were generated as described previously (Wu et al., 1990). Briefly, the interior of a gastrula from stage 6/7 embryo was extracted with a glass micropipette and dispersed into a drop of culture medium (∼ 40 μm) on an uncoated coverslip. The culture medium contained 80% Drosophila Schneider’s insect medium (S0146-100ML, Merck KGaA, Darmstadt, Germany) and 20% fetal bovine serum (F2442-100ML, Merck KGaA, Darmstadt, Germany), with the addition of 50 μg/ml streptomycin, 50 U/ml penicillin (all from Sigma, St. Louis, MO, USA) and cytochalasin B (CCB; 2 μg/ml; Sigma, St. Louis, MO, USA). CCB was removed by washing with CCB-free medium 1 d after plating. All cultures were grown in a humidified chamber.

To label Brp::SNAP, cells were incubated in fluorescent SNAP substrate diluted in the culture medium (1:5000, SNAP-Cell 647-SiR) for half an hour at RT. Subsequently, the same cells were labeled by the SNAP substrate to minimize the crosstalk. After the incubation, cells were washed with the culture medium and subjected to confocal scanning with Leica TCS SP8 microscopy equipped with a 40x oil immersion objective (HC PL APO 40x/1.30 Oil PH3 CS2, Leica). Acquired images were then processed with the LIGHTNING software.

Behavioral assays

The experimental protocols for appetitive learning experiment (Figure 9C) were as described previously (Ichinose and Tanimoto., 2016). For appetitive conditioning, a group of approximately 50 flies in a training tube alternately received 3-octanol (3OCT; Merck) and 4-methylcyclohexanol (4MCH; Sigma-Aldrich), for 1 minute in a constant air flow with or without reward with an interval of 1 minute between the two odor presentations. These odorants were diluted to 1.2% and 2% with paraffin wax (Sigma-Aldrich), respectively. Dried 2 M sucrose (Sigma-Aldrich) on a piece of filter paper was used as the reward. Flies were starved in the presence of wet tissue paper for 24 hours before appetitive conditioning. For testing, flies were given a choice between the odor paired with reward (Conditioned stimulus, CS+) and another one unpaired (CS-). Their distribution in the plexiglass T-maze was video-recorded one frame per second for 2 minutes with the CMOS cameras (GS3-U3-51S5M, FLIR) under infrared LED illumination (Ichinose & Tanimoto, 2016). Flies were forced to stay on the floor by applying Fluon (Insect-a-Slip PTFE30, BioQuip Products) on the side and top of T-maze arms.

Learning index was then calculated by counting the number of flies in each arm using ImageJ macro as described previously (Ichinose and Tanimoto, 2016) with the following formula (Tempel et al., 1983): where #CS+ and #CS− imply the number of flies in the CS+ and the CS-arms, respectively. Learning index was calculated for every second and was averaged for the last 60 s of the 120 s test session. An average of a pair of reciprocally trained groups was used as a single data point.

In aversive associative learning experiment (Figure 2 – figure supplement 1), flies were conditioned with twelve 1.5 second pulses of 90V electric shock paired with either 4MCH or 3OCT (CS+) for 1 minute. After 1 minute of interval, another odor (CS-) was presented without electric shock for 1 minute. Flies were tested immediately after training. In the test, flies were allowed to choose between the CS+ and CS-odors in a T-maze. The number of flies in each side of the arm (#CS+ and #CS-) after 2 minutes was used to calculate learning index as described above.

Statistics

For multiple comparison (Figure 2 – figure supplement 1, Figure 8C, 8F and Figure 8 – figure supplement 1 and Figure 9C), statistics were performed by Prism5 (GraphPad). Data were always first tested for normality (Shapiro-Wilk test) and for homoscedasticity (Spearman’s test or Bartlett’s test). If these assumptions were not violated, parametric tests (one- or two-way ANOVA, followed by Dunnett’s test or t-test with Bonferroni correction) were applied. If data did not suffice the assumptions for parametric tests, nonparametric tests (Kruskal-Wallis, followed by Dunn’s post hoc pairwise test) were performed. For comparison of two groups, Student’s t test (Figure 3D and Figure 3 – figure supplement 1) or Mann-Whitney U test (Figure 7E) were performed.

Bars and error bars represent means and SEM, respectively, in all figures. For all figures, significance corresponds to the following symbols: * p < 0.05; ** p < 0.01; *** p < 0.001; ns: not significant p>0.05.